# Week 4: Data Frames and Plotting {#sec-week4}

```{r}
#| include: false
set.seed(7432)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, out.width = "70%", fig.align = "center")
```

```{python}
#| include: false
import numpy as np
rng = np.random.default_rng(7432)   # the Python companion of set.seed(7432)
import plotnine
plotnine.options.figure_size = (6, 4)
```

Three weeks of machinery: vectors and matrices (@sec-week1), loops (@sec-week2), functions
(@sec-week3). This week we point all of it at what statisticians actually spend their days on:
tables of data. We will learn how to store data in *data frames*, manipulate them with `dplyr`
(in Python: `pandas`), and, at last, plot them properly with `ggplot2` (in Python:
`plotnine`). The quick-and-dirty `plot()` and `plt.plot()` calls from the previous weeks retire
this week.

## The tidyverse (and friends in Python)

Modern data work in R happens largely inside the **tidyverse**, a collection of packages that
share a common design: `dplyr` for manipulation, `readr` for import/export, `tibble` for nicer
data frames, `ggplot2` for graphics, and `purrr`, which you already met in @sec-week3.

**Note for Python:** Python spreads the same jobs over separate packages: `pandas` covers
what `dplyr`, `readr` and `tibble` do, and `plotnine` is a deliberate port of `ggplot2` (same
grammar, nearly the same code). Together with `numpy` they are this week's toolkit, installed
and imported alongside the R column below.

Install them once (already done on university machines). The R line runs in the R console; the
Python line is a shell command, so it runs in a terminal, not in the Python console:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
#| eval: false
install.packages("tidyverse")
```

</div>

<div>

**Python**

```{python}
#| eval: false
python -m pip install pandas plotnine
```

</div>
:::

then load them at the top of every script:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
#| message: false
library(tidyverse)
```

</div>

<div>

**Python**

```{python}
# several packages instead of one: import them together, at the top
import numpy as np
import pandas as pd
from plotnine import (
    ggplot, aes, geom_point, geom_smooth, geom_histogram, geom_boxplot,
    geom_line, geom_hline, facet_wrap, labs, theme_minimal
)
```

</div>
:::

Installing happens once per computer; loading happens once per session: a distinction everyone
mixes up exactly once.

## Data frames {#sec-week4-df}

A **data frame** is a table: each column is a named vector, and different columns may hold
different types (numeric, character, logical). This is the crucial difference from the matrices
of @sec-week1, where every entry had to share one type. In statistics-speak, columns are
**variables** and rows are **observations**.

### Building one

Here is a small dataset of plant-based milks and their nutritional content per 250 ml glass:

| type   | calories | protein | fibre | carbs | sugars | fat |
|--------|----------|---------|-------|-------|--------|-----|
| Almond | 60       | 1       | 1     | 8     | 7      | 2.5 |
| Soy    | 100      | 7       | 2     | 10    | NA     | 4   |
| Oat    | 120      | 3       | NA    | 20    | 19     | 2.5 |
| Rice   | 130      | 1       | NA    | 25    | NA     | 2.5 |

Note the `NA`s: real data is rarely complete, and R has a dedicated value for "not available"
(in Python, `pandas` marks missing numbers with the floating-point `NaN`, written `np.nan`).
We build the data frame column by column with `data.frame()` (in Python, from a dictionary
mapping column names to lists):

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
# each argument is one column (a vector); all must have the same length
milk <- data.frame(
  type     = c("Almond", "Soy", "Oat", "Rice"),
  calories = c(60, 100, 120, 130),
  protein  = c(1, 7, 3, 1),
  fibre    = c(1, 2, NA, NA),
  carbs    = c(8, 10, 20, 25),
  sugars   = c(7, NA, 19, NA),
  fat      = c(2.5, 4, 2.5, 2.5)
)
milk
```

</div>

<div>

**Python**

```{python}
# each entry is one column (a list); all must have the same length
milk = pd.DataFrame({
    "type":     ["Almond", "Soy", "Oat", "Rice"],
    "calories": [60, 100, 120, 130],
    "protein":  [1, 7, 3, 1],
    "fibre":    [1, 2, np.nan, np.nan],
    "carbs":    [8, 10, 20, 25],
    "sugars":   [7, np.nan, 19, np.nan],
    "fat":      [2.5, 4, 2.5, 2.5]
})
milk
```

</div>
:::

### Columns, rows and cells

Single columns come out with the dollar syntax in R and single square brackets in Python,
returning a plain vector (in pandas: a `Series`):

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
milk$calories
```

</div>

<div>

**Python**

```{python}
milk["calories"]
```

</div>
:::

In R, square brackets with a column name return instead a *smaller data frame*. pandas does
the same when you double the brackets:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
milk["calories"]           # still a data frame
milk[c("type", "protein")] # two columns, still a data frame
```

</div>

<div>

**Python**

```{python}
milk[["calories"]]          # double brackets: still a data frame
milk[["type", "protein"]]   # two columns, still a data frame
```

</div>
:::

**Note for Python:** the single/double bracket split is worth committing to memory:
`milk["calories"]` is a `Series` (the vector-like object), `milk[["calories"]]` is a one-column
`DataFrame`. The attribute shorthand `milk.calories` also works, but only for names that are
valid Python identifiers. Brackets are the safe habit.

For rows (and row/column combinations) we index in two dimensions, exactly as we did with
matrices: `df[rows, columns]`, with an empty slot meaning "everything":

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
milk[2, ]                        # second row, all columns
milk[1:2, c("type", "calories")] # first two rows, two columns
```

</div>

<div>

**Python**

```{python}
milk.iloc[[1]]                       # second row: position 1 (Python counts from 0!)
milk.loc[0:1, ["type", "calories"]]  # first two rows, two columns
```

</div>
:::

**Note for Python:** pandas splits two-dimensional indexing over two *indexers*: `.iloc`
takes integer **positions** (0-based, end excluded: the rules of @sec-week1), while `.loc`
takes **labels** (here the row labels happen to be the integers 0, 1, 2, ...) and includes
*both* ends, which is why `0:1` above returns two rows. Also, `milk.iloc[1]` without the inner
brackets squeezes the row into a `Series`; the doubled brackets keep it a one-row `DataFrame`.

The comparison operators of @sec-week2 work here too. A logical vector selects rows:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
# rows where calories exceed 100
milk[milk$calories > 100, ]
```

</div>

<div>

**Python**

```{python}
# rows where calories exceed 100
milk[milk["calories"] > 100]
```

</div>
:::

New columns are created by assigning to a name that does not exist yet:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
# what fraction of the carbs is sugar?
milk$sugar_frac <- milk$sugars / milk$carbs
milk
```

</div>

<div>

**Python**

```{python}
# what fraction of the carbs is sugar?
milk["sugar_frac"] = milk["sugars"] / milk["carbs"]
milk
```

</div>
:::

### Getting to know a data frame

For a data frame you did not build yourself, four functions tell you almost everything:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
nrow(milk)     # number of rows (ncol() and dim() also exist)
str(milk)      # structure: column names, types, first values
summary(milk)  # summary statistics per column, NAs included
head(milk, 2)  # first few rows (useful for big data)
```

</div>

<div>

**Python**

```{python}
len(milk)        # number of rows (milk.shape gives rows and columns)
milk.info()      # structure: column names, types, non-missing counts
milk.describe()  # summary statistics per numeric column
milk.head(2)     # first few rows (useful for big data)
```

</div>
:::

`summary()` counts the `NA`s for us, and those `NA`s are contagious. Any computation touching a
missing value returns `NA`, unless you explicitly ask R to drop them:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
mean(milk$sugars)               # NA: R refuses to guess
mean(milk$sugars, na.rm = TRUE) # mean of the observed values
```

</div>

<div>

**Python**

```{python}
milk["sugars"].mean()               # pandas skips NaN by default!
milk["sugars"].mean(skipna=False)   # the R-style refusal is opt-in
```

</div>
:::

**Note for Python:** the two languages picked *opposite* defaults: R propagates `NA`
until you say `na.rm = TRUE`; pandas silently skips `NaN` until you say `skipna=False`. Neither
is wrong, but know which one you are getting: a summary that quietly ignores missing values
can hide a data problem R would have shouted about.

::: {.callout-tip title="Exercise 4.1: A data frame of distributions"}
Build a data frame `dists` describing five distributions, with columns:

- `name`: `"Bernoulli"`, `"Poisson"`, `"Exponential"`, `"Gaussian"`, `"Uniform"`;
- `n_params`: the number of parameters of each: 1, 1, 1, 2, 2;
- `discrete`: `TRUE, TRUE, FALSE, FALSE, FALSE`.

Then:

a. Extract the `name` column as a vector.
b. Using logical indexing, extract the rows of the discrete distributions. (Hint: `discrete` is
   already a logical vector, so no comparison is needed.)
c. Add a logical column `symmetric` using the dollar syntax (Python: bracket assignment). Only
   the Gaussian and the Uniform are symmetric.
d. Check your work with `str()` and `dim()` (Python: `.info()` and `.shape`).
:::

## Tibbles (and the pandas index) {#sec-week4-tibbles}

The tidyverse ships a modernised data frame called a **tibble**. It *is* a data frame
(everything above applies) with two behavioural upgrades. First, printing: a tibble shows ten
rows, only the columns that fit, and the type of each. A classic data frame would happily dump
every last row of a 50,000-row dataset into your console. Second, no partial matching. Watch
this data.frame quirk:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
milk$cal   # base data frame: silently matches "calories"!
```

</div>

<div>

**Python**

```{python}
#| error: true
milk["cal"]   # pandas never guesses: unknown column, immediate error
```

</div>
:::

That is a bug waiting to happen: mistype a column name and R may "helpfully" return a different
column. A tibble warns you and returns `NULL`:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
as_tibble(milk)$cal
```

</div>

<div>

**Python**

```{python}
print(milk.get("cal"))   # the forgiving lookup: None (Python's NULL)
```

</div>
:::

**Note for Python:** there is no tibble to reach for because there is no quirk to fix:
pandas has exactly **one** table type, the `DataFrame` you have been using all along, and it
never partial-matches (as the `KeyError` above shows; `.get()` is the polite version). The
genuinely new thing to watch is the **index**: every `DataFrame` carries row labels (the
unnamed left column in the printouts), which stick to their rows through filtering and sorting.
It is pandas' (heavily used) version of R's (rarely used) row names; when it gets in your way,
`.reset_index()` demotes it to an ordinary column, and `index=False` keeps it out of your CSV
files. Both will appear below.

You create tibbles directly with `tibble()` (same syntax as `data.frame()`, with the bonus that
later columns can use earlier ones), or convert with `as_tibble()`. From now on we work with
tibbles, and we need a bigger dataset to play with: `mtcars`, a built-in table of 32 cars from a
1974 issue of *Motor Trend* (in Python it ships with plotnine, as `plotnine.data.mtcars`). The
variables we will use most: `mpg` (miles per gallon; higher means more fuel-efficient), `cyl`
(cylinders), `hp` (horsepower), `wt` (weight, 1000 lbs), `qsec` (quarter-mile time in seconds),
`am` (transmission: 0 automatic, 1 manual).

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
head(mtcars, 3)
```

</div>

<div>

**Python**

```{python}
from plotnine.data import mtcars
mtcars.head(3)
```

</div>
:::

One wrinkle: `mtcars` stores the car names as *row names*, a base-R feature tibbles deliberately
drop. `as_tibble()` can rescue them into a proper column:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
cars <- as_tibble(mtcars, rownames = "model")
cars
```

</div>

<div>

**Python**

```{python}
# plotnine's copy already keeps the names in a column; just rename it
cars = mtcars.rename(columns={"name": "model"})
cars
```

</div>
:::

**Note for Python:** pandas' index (see the note above) is exactly the row-names idea, and
`mtcars.set_index("name")` would recreate R's original layout. But, as in R, the tidy habit is
to keep data in ordinary columns.

## Reading and writing data {#sec-week4-io}

Data usually lives in files, and the lingua franca of data files is the CSV (comma-separated
values). The `readr` package (in the tidyverse) provides `write_csv()` and `read_csv()`:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
# write the milk data next to this script...
write_csv(milk, "week4_milk.csv")

# ...and read it back
milk2 <- read_csv("week4_milk.csv")
milk2
```

</div>

<div>

**Python**

```{python}
# write the milk data next to this script...
milk.to_csv("week4_milk_py.csv", index=False)

# ...and read it back
milk2 = pd.read_csv("week4_milk_py.csv")
milk2
```

</div>
:::

`read_csv()` reads the file, guesses each column's type (it prints its guesses: read them, they
catch import bugs early), and returns a tibble. Base R has `read.csv()` (dot, not underscore),
which does the same job into a plain data frame, more slowly on large files. Recall from
@sec-week1 that a bare filename like `"week4_milk.csv"` is relative to your working directory:
if the file cannot be found, `getwd()` (Python: `os.getcwd()`) is the first thing to check.

**Note for Python:** two things in the chunk above. First, the filename: throughout these
notes the Python column writes `_py` files (`week4_milk_py.csv`) so the two languages never
overwrite each other's output. Second, `index=False`: without it pandas saves the index as an
anonymous extra column, which comes back to haunt you at reading time. `pd.read_csv()` guesses
column types just like `readr`, but silently; run `.info()` on an import when it matters.

::: {.callout-tip title="Exercise 4.2: A CSV round trip"}
Using the `dists` data frame from Exercise 4.1:

a. Write it to `"week4_dists.csv"` with `write_csv()` (Python: to `"week4_dists_py.csv"` with
   `.to_csv(..., index=False)`).
b. Find the file and open it with a text editor (or click it in the RStudio Files pane and choose
   *View File*). What does the raw format look like?
c. Read it back with `read_csv()` (Python: `pd.read_csv()`) and compare `str()` (Python:
   `.info()`) of the original and the copy. Did the logical column survive the trip?
d. Read it once more with base `read.csv()`. What differs? (R only: Python has just the one
   CSV reader.)
:::

## The pipe `|>` (and method chaining) {#sec-week4-pipe}

Data analysis is a sequence of steps, and code reads best when it is written in the order the
steps happen. The **pipe** operator `|>` takes the value on its left and passes it as the *first
argument* of the function on its right: programming's version of function composition
$(f_2 \circ f_1)(x)$, which we met in @sec-week3. So to compute $\ln(1 + 2 + \dots + 10)$:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
x <- 1:10
x |> sum() |> log()   # same as log(sum(x))
```

</div>

<div>

**Python**

```{python}
x = np.arange(1, 11)
np.log(x.sum())       # no pipe in Python: nest the calls, or chain methods
```

</div>
:::

Read it left to right: "take `x`, sum it, take the log". Further arguments go in their usual
place, after the piped-in first one:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
x |> sum() |> log(base = 10)   # same as log(sum(x), base = 10)
```

</div>

<div>

**Python**

```{python}
np.log10(x.sum())     # numpy's dedicated base-10 logarithm
```

</div>
:::

In the wild you will also meet `%>%`, the older pipe from the `magrittr` package. For our
purposes it does the same thing, but write the native `|>`.

The pipe matters this week because every `dplyr` verb takes a data frame as its first argument
and returns a data frame: they are *built* to be piped.

**Note for Python:** Python has no pipe operator; the idiom that plays its role is
**method chaining**. Every pandas verb below is a *method* that returns a new `DataFrame`, so a
pipeline reads left to right as `cars.query(...).assign(...).sort_values(...)`: dots instead
of pipes. Wrap a chain in parentheses to split it over several lines, one verb per line; you
will see that layout constantly from here on.

## Data manipulation with dplyr (and pandas) {#sec-week4-dplyr}

`dplyr` provides one verb per manipulation you would want to do to a table:

| Verb          | Action                                             | pandas counterpart        |
|---------------|----------------------------------------------------|---------------------------|
| `filter()`    | keep a subset of the **rows**                      | `.query()` / boolean mask |
| `select()`    | keep a subset of the **columns**                   | `[[...]]` double brackets |
| `mutate()`    | add new columns computed from existing ones        | `.assign()`               |
| `arrange()`   | reorder the rows                                   | `.sort_values()`          |
| `summarise()` | collapse rows down to summary statistics           | `.agg()`                  |

Inside a verb you refer to columns by bare name (no `cars$` prefix needed); `dplyr` knows to
look inside the data frame first. (One side effect of loading the tidyverse: its `filter()`
masks the base function of the same name. If `filter()` ever behaves strangely, be explicit and
call `dplyr::filter()`.)

**Note for Python:** pandas gives you the bare-name comfort only inside `.query()`, whose
condition is a *string* evaluated inside the data frame. Elsewhere columns are spelled
`cars["mpg"]`. Inside a chain, where the intermediate data frame has no name, `.assign()`
accepts functions of it (`lambda df: ...`). Both appear below.

### filter and select

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
# rows: frugal cars only
cars |> filter(mpg > 25)

# conditions combine with the logical operators of week 2
cars |> filter(cyl == 4 & hp > 90)
```

</div>

<div>

**Python**

```{python}
# rows: frugal cars only
cars.query("mpg > 25")

# conditions combine with the logical operators of week 2
cars.query("cyl == 4 & hp > 90")
```

</div>
:::

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
# columns: just the ones we care about
cars |> select(model, mpg, hp) |> head(3)

# or everything except some
cars |> select(-vs, -am) |> head(3)
```

</div>

<div>

**Python**

```{python}
# columns: just the ones we care about
cars[["model", "mpg", "hp"]].head(3)

# or everything except some
cars.drop(columns=["vs", "am"]).head(3)
```

</div>
:::

### mutate and arrange

`mutate()` adds columns. Let's add fuel efficiency in sensible units, and, while we are at it,
turn the 0/1 codes into labelled categorical variables. R's type for categorical data is the
**factor**: a vector with a fixed set of levels and readable labels, which `summary()` and
`ggplot2` treat as labelled groups (discrete colour scales, one box or line per level) rather
than as numbers on a continuous axis.

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
cars <- cars |>
  mutate(
    kpl = mpg * 0.425,                                # kilometres per litre
    am  = factor(am, labels = c("automatic", "manual")),
    vs  = factor(vs, labels = c("V-shaped", "straight"))
  )
```

</div>

<div>

**Python**

```{python}
cars = cars.assign(
    kpl = cars["mpg"] * 0.425,                        # kilometres per litre
    am  = pd.Categorical(cars["am"].map({0: "automatic", 1: "manual"})),
    vs  = pd.Categorical(cars["vs"].map({0: "V-shaped", 1: "straight"}))
)
```

</div>
:::

**Note for Python:** the factor's pandas counterpart is the **categorical** dtype:
`.map()` swaps the 0/1 codes for readable labels, and `pd.Categorical()` marks the result as
categorical, which plotnine then treats exactly as ggplot2 treats factors: discrete colour
scales, one box or line per level.

`arrange()` sorts rows, ascending by default, descending inside `desc()` (pandas:
`ascending=False`). And because every verb pipes into the next, a whole analysis becomes one
readable chain:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
# the most frugal 4-cylinder cars
cars |>
  filter(cyl == 4) |>
  arrange(desc(kpl)) |>
  select(model, kpl, hp) |>
  head(3)
```

</div>

<div>

**Python**

```{python}
# the most frugal 4-cylinder cars
(
  cars
  .query("cyl == 4")
  .sort_values("kpl", ascending=False)
  [["model", "kpl", "hp"]]
  .head(3)
)
```

</div>
:::

::: {.callout-tip title="Exercise 4.3: dplyr drills"}
Working on `cars` (the `mtcars` table with the labelled categorical columns from above):

a. Find all manual cars that do more than 25 miles per gallon.
b. Add a column `efficiency = mpg / hp`, miles per gallon per horsepower.
c. Show the `model`, `mpg`, `hp` and `efficiency` of the five most efficient cars.
d. Now do (b) and (c) as a single pipeline, starting from `cars` and using each verb at most
   once.
:::

### group_by and summarise {#sec-week4-groupby}

`summarise()` collapses a data frame to summary statistics: `n()` counts the rows, and any
vector-to-number function (`mean`, `sd`, `median`, `min`, ...) works:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
cars |> summarise(n = n(), mean_mpg = mean(mpg), sd_mpg = sd(mpg))
```

</div>

<div>

**Python**

```{python}
# ungrouped: pick the column, name the summaries
cars["mpg"].agg(n = "size", mean_mpg = "mean", sd_mpg = "std")
```

</div>
:::

On its own this is a glorified `mean()`. The power move is `group_by()`: it splits the data
frame into groups by the values of one or more columns, so that `summarise()` produces *one row
per group*. Group means and group spreads are the first step of nearly every statistical
analysis, and here they cost one pipeline:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
cars |>
  group_by(am) |>
  summarise(n = n(), mean_mpg = mean(mpg), sd_mpg = sd(mpg))
```

</div>

<div>

**Python**

```{python}
(
  cars
  .groupby("am", observed=True)
  .agg(n        = ("mpg", "size"),
       mean_mpg = ("mpg", "mean"),
       sd_mpg   = ("mpg", "std"))
  .reset_index()
)
```

</div>
:::

Manual cars in this dataset do, on average, about 7 more miles per gallon, though with 1974
technology, the causal story is mostly "small cheap cars had manual gearboxes". Grouping by two
variables gives one row per combination:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
#| message: false
cars |>
  group_by(am, cyl) |>
  summarise(n = n(), mean_mpg = mean(mpg))
```

</div>

<div>

**Python**

```{python}
(
  cars
  .groupby(["am", "cyl"], observed=True)
  .agg(n = ("mpg", "size"), mean_mpg = ("mpg", "mean"))
  .reset_index()
)
```

</div>
:::

**Note for Python:** each `.agg()` pair reads `new_column = (column, function)`: the
counterpart of `summarise()`'s `new_column = f(column)`. `groupby()` moves the grouping values
into the index, so we finish with `.reset_index()` to get ordinary columns back; and
`observed=True` keeps only group combinations that actually occur, which is what `summarise()`
does (it also silences a pandas warning about its changing default).

::: {.callout-tip title="Exercise 4.4: Group statistics on mtcars"}
a. Group `cars` by number of cylinders and compute, per group: the number of cars, the mean and
   standard deviation of `mpg`, and the mean of `hp`.
b. Group by transmission *and* cylinders, computing group sizes and mean `mpg`.
c. Compare (b) with the transmission-only summary in the notes: within each cylinder class, is
   the manual advantage still around 7 mpg? What does that tell you about comparing group means
   without controlling for other variables?
:::

### Joining data frames

Real analyses often need to combine two tables that share a common column (a *key*). Suppose we
find supermarket prices for some of our milks:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
prices <- tibble(
  type  = c("Almond", "Oat", "Soy"),
  price = c(1.80, 1.50, 1.40)   # £ per litre
)

# match rows of `milk` with rows of `prices` by the shared column `type`
left_join(milk, prices, by = "type")
```

</div>

<div>

**Python**

```{python}
prices = pd.DataFrame({
    "type":  ["Almond", "Oat", "Soy"],
    "price": [1.80, 1.50, 1.40]   # £ per litre
})

# match rows of `milk` with rows of `prices` by the shared column `type`
milk.merge(prices, on="type", how="left")
```

</div>
:::

`left_join()` keeps every row of its first argument and attaches the matching information from
the second; rows with no match (poor Rice) get `NA` (`NaN` in pandas). There is a whole family (`full_join()`,
`inner_join()`, `right_join()`) differing only in which unmatched rows survive. (pandas folds
the whole family into `.merge()`'s `how` argument: `"left"`, `"outer"`, `"inner"`, `"right"`.)

## Plotting with ggplot2 (and plotnine) {#sec-week4-ggplot}

`ggplot2` implements the **grammar of graphics**: instead of memorising one function per chart
type, you describe a plot with three ingredients:

- the **data**: a data frame;
- the **aesthetic mapping** `aes()`: which variables map to which visual channels (x-position,
  y-position, colour, fill, shape);
- the **geometries** `geom_*()`: how the mapped data is drawn (points, lines, bars, boxes).

Layers are glued together with `+` (a historical quirk: ggplot2 predates the pipe). Swap the
geometry and the same data-to-aesthetics mapping becomes a different plot; that is the whole
grammar. Easier done than said:

**Note for Python:** `plotnine` is a faithful port of `ggplot2`: same grammar, same
function names, same `+`. Two mechanical differences only: column names inside `aes()` are
quoted strings (`aes(x="wt", y="mpg")`) and a plot spread over several lines needs wrapping
parentheses (plus `.show()` to display it from a script). Both spellings of `colour`/`color`
work, in either language.

### Scatter plots

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
ggplot(cars, aes(x = wt, y = mpg)) +
  geom_point()
```

</div>

<div>

**Python**

```{python}
(
  ggplot(cars, aes(x="wt", y="mpg")) +
  geom_point()
).show()
```

</div>
:::

Dissecting it: `ggplot(cars, ...)` starts a plot from the `cars` data; `aes(x = wt, y = mpg)`
maps weight to the x-axis and fuel consumption to the y-axis; `+ geom_point()` draws one point
per row. Heavier cars burn more fuel, a relationship crying out for a line through it:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
#| message: false
ggplot(cars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

</div>

<div>

**Python**

```{python}
(
  ggplot(cars, aes(x="wt", y="mpg")) +
  geom_point() +
  geom_smooth(method="lm", se=False)
).show()
```

</div>
:::

`geom_smooth(method = "lm")` overlays the least-squares regression line: the line minimising
the sum of squared vertical distances to the points, the same fit you would get from R's
`lm(mpg ~ wt, data = cars)` (in Python, from `statsmodels`, which is exactly what plotnine's
`method="lm"` calls under the hood). Fitting models by minimising a loss is precisely the
business of @sec-week5.

To bring in a third variable, map it to a third aesthetic:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
ggplot(cars, aes(x = wt, y = mpg, colour = am)) +
  geom_point()
```

</div>

<div>

**Python**

```{python}
(
  ggplot(cars, aes(x="wt", y="mpg", colour="am")) +
  geom_point()
).show()
```

</div>
:::

One extra word in `aes()` bought us a coloured grouping *and* a legend.

### Mapping vs setting aesthetics

There are two ways to control colour, and mixing them up is the most common ggplot bug in
existence. **Mapping** (inside `aes()`) ties an aesthetic to a *variable*; **setting** (outside
`aes()`, inside the geom) fixes it to a *constant*:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
# setting: every point steelblue, no legend (a constant, so OUTSIDE aes())
ggplot(cars, aes(x = wt, y = mpg)) +
  geom_point(colour = "steelblue")
```

</div>

<div>

**Python**

```{python}
# setting: every point steelblue, no legend (a constant, so OUTSIDE aes())
(
  ggplot(cars, aes(x="wt", y="mpg")) +
  geom_point(colour="steelblue")
).show()
```

</div>
:::

Put a constant *inside* `aes()` and ggplot does something legal but absurd: it invents a
variable whose every value is the string `"blue"`, and dutifully assigns it the first default
colour, which is red:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
# the classic mistake: "blue" is not a variable
ggplot(cars, aes(x = wt, y = mpg, colour = "blue")) +
  geom_point()
```

</div>

<div>

**Python**

```{python}
# the classic mistake: "blue" is not a variable
(
  ggplot(cars, aes(x="wt", y="mpg", colour='"blue"')) +
  geom_point()
).show()
```

</div>
:::

If your plot sports a legend titled `"blue"` full of red points, you now know why.

**Note for Python:** mind the doubled quotes in `aes(colour='"blue"')`. plotnine
evaluates `aes()` strings as expressions, so the inner quotes are what make `"blue"` a constant
string: a plain `aes(colour="blue")` would go looking for a *column* called `blue` and raise
an error instead of drawing red points. Different failure, same moral: constants do not belong
in `aes()`.

::: {.callout-tip title="Exercise 4.5: Reproduce this scatter plot"}
Build, from `cars`, the following plot (described in words: your job is to translate it into
one ggplot call):

- a scatter plot with gross horsepower `hp` on the x-axis and quarter-mile time `qsec` on the
  y-axis;
- points coloured by transmission (`am`);
- one least-squares line per transmission group, without the confidence band.

You should find that mapping `colour` automatically makes `geom_smooth()` fit one line per
group. Does more horsepower mean a faster quarter mile? Does the relationship differ between
transmissions?
:::

### Histograms and boxplots

The distribution of a single variable needs only an `x` mapping; `geom_histogram()` bins it and
draws bar heights:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
ggplot(cars, aes(x = mpg)) +
  geom_histogram(binwidth = 2)
```

</div>

<div>

**Python**

```{python}
(
  ggplot(cars, aes(x="mpg")) +
  geom_histogram(binwidth=2)
).show()
```

</div>
:::

The `binwidth` is your choice, and an important one: too narrow is noise, too wide smooths away
every feature. Always try a few values.

For comparing a distribution *across groups*, the boxplot is the tool: a box from the first to
the third quartile with the median inside, whiskers spanning the rest of the data, and any
outliers as lone points. Map the group to `x` and the variable to `y`; boxes and bars are filled
shapes, so their interior colour is the `fill` aesthetic (`colour` is just the outline):

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
ggplot(cars, aes(x = am, y = mpg, fill = am)) +
  geom_boxplot()
```

</div>

<div>

**Python**

```{python}
(
  ggplot(cars, aes(x="am", y="mpg", fill="am")) +
  geom_boxplot()
).show()
```

</div>
:::

The medians echo our `group_by()` summary (manual cars sit visibly higher), but the boxplot
also shows the overlap and spread that a table of means quietly hides.

::: {.callout-tip title="Exercise 4.6: Petal boxplots"}
The built-in `iris` data frame records sepal and petal measurements for 150 flowers of three
species. (Python: `iris` ships with R, not Python. Download `iris.csv` from Moodle
into your working directory and load it with `iris = pd.read_csv("iris.csv")`; the R-style
column names such as `Petal.Length` carry over, and plotnine accepts them.)

a. Draw boxplots of `Petal.Length` by `Species`, with each box filled in a different colour.
b. Which species is clearly separated from the others? Would a single overall histogram of
   `Petal.Length` have shown you this?
c. Change `fill` from a mapping to a setting: fill *all* boxes `"orchid"`. Where does
   `fill = "orchid"` have to go?
:::

### Line plots

`geom_line()` connects observations in x-order: the right geometry when x is time, iteration
count, or anything sequential. Let's watch the law of large numbers in action: simulate 10000
fair-coin flips (`rbinom(10000, size = 1, prob = 0.5)` in R, `rng.binomial(1, 0.5, size=10000)`
in Python; both draw 10000 independent 0/1 values) and plot the running proportion of heads.
Since randomness is involved, we set the seed so that each "random" figure is exactly
reproducible, though not the *same* figure: R's and numpy's generators produce different
streams, so the two columns show two different, equally fair, coin sequences:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
set.seed(7432)
lln <- tibble(
  n            = 1:10000,
  running_mean = cumsum(rbinom(10000, size = 1, prob = 0.5)) / n
)

ggplot(lln, aes(x = n, y = running_mean)) +
  geom_line() +
  geom_hline(yintercept = 0.5, linetype = "dashed")
```

</div>

<div>

**Python**

```{python}
rng = np.random.default_rng(7432)
n = np.arange(1, 10001)
lln = pd.DataFrame({
    "n":            n,
    "running_mean": np.cumsum(rng.binomial(1, 0.5, size=10000)) / n
})

(
  ggplot(lln, aes(x="n", y="running_mean")) +
  geom_line() +
  geom_hline(yintercept=0.5, linetype="dashed")
).show()
```

</div>
:::

In both runs the wild early swings settle onto the dashed line at $1/2$: the sample mean
converging to the expectation, slowly, as the fluctuations only shrink like $1/\sqrt{n}$.
(`geom_hline()` draws a horizontal reference line; note how it stacks onto the plot as just
another layer.) In @sec-week5 the very same plot template (value against iteration) will
track optimisation algorithms as they converge.

**Note for Python:** watch the argument names: they swap roles across the languages.
R's `rbinom(10000, size = 1, prob = 0.5)` asks for 10000 draws of `size = 1` trial each;
numpy's `rng.binomial(1, 0.5, size=10000)` says the same thing, but now `size` counts the
*draws*. Copying arguments across languages by name, without reading, is how "10000 flips of
one coin" silently becomes "one flip of 10000 coins".

### Facets

Colour handles a third variable; a fourth would get crowded. **Faceting** splits one plot into a
panel per value of a variable, with shared axes so panels compare fairly:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
ggplot(cars, aes(x = wt, y = mpg, colour = am)) +
  geom_point() +
  facet_wrap(~ cyl)
```

</div>

<div>

**Python**

```{python}
(
  ggplot(cars, aes(x="wt", y="mpg", colour="am")) +
  geom_point() +
  facet_wrap("~cyl")
).show()
```

</div>
:::

`facet_wrap(~ cyl)` means "one panel per value of `cyl`" (the `~` is R's formula syntax; you
will meet it again in regression). Arrange the panels with `nrow` or `ncol`, e.g.
`facet_wrap(~ cyl, ncol = 1)` stacks them vertically, which is ideal for comparing
distributions along a shared x-axis.

**Note for Python:** plotnine takes the formula as a string (`facet_wrap("~cyl")`, or
just `facet_wrap("cyl")`), with the same `nrow`/`ncol` arguments.

::: {.callout-tip title="Exercise 4.7: Facet the histograms"}
Back to `iris`: draw histograms of `Petal.Length` (pick a sensible `binwidth`), faceted by
`Species`, with the three panels stacked in a single column. Why does stacking them vertically
(rather than side by side) make the species easier to compare? How does this figure relate to
the boxplots of Exercise 4.6?
:::

### Labels, themes and saving

Everything so far was exploratory: quick plots for you. A figure for a report needs proper
labels (nobody downstream knows what `wt` means) and a cleaner look. `labs()` names every
mapped aesthetic plus the title, and `theme_minimal()` swaps the grey default for something
publication-friendly (we will use it from now on):

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
p <- ggplot(cars, aes(x = wt, y = mpg, colour = am)) +
  geom_point() +
  labs(
    title  = "Lighter cars go further on a gallon",
    x      = "Weight (1000 lbs)",
    y      = "Miles per US gallon",
    colour = "Transmission"
  ) +
  theme_minimal()
p
```

</div>

<div>

**Python**

```{python}
p = (
  ggplot(cars, aes(x="wt", y="mpg", colour="am")) +
  geom_point() +
  labs(
      title  = "Lighter cars go further on a gallon",
      x      = "Weight (1000 lbs)",
      y      = "Miles per US gallon",
      colour = "Transmission"
  ) +
  theme_minimal()
)
p.show()
```

</div>
:::

Note we stored the plot in a variable: plots are ordinary objects in both languages, so you
can print them, add more layers later (`p + facet_wrap(~ cyl)`), or save them to a file with
`ggsave()` (in plotnine, the plot's `.save()` method):

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
ggsave("week4_mpg_wt.png", p, width = 6, height = 4)  # inches
```

</div>

<div>

**Python**

```{python}
p.save("week4_mpg_wt_py.png", width=6, height=4, verbose=False)  # inches
```

</div>
:::

Change the extension to get a different format (`.pdf` scales beautifully in documents). If a
figure looks wrong in your report, resize it here, in code, never by dragging its corner in a
word processor.

## Challenges

::: {.callout-important title="Challenge 4.1: The full pipeline"}
Time to run the whole week end to end: file → table → wrangle → summarise → figure → file.
First create the raw data file by running:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
write_csv(as_tibble(mtcars, rownames = "model"), "week4_cars.csv")
```

</div>

<div>

**Python**

```{python}
mtcars.rename(columns={"name": "model"}).to_csv("week4_cars_py.csv", index=False)
```

</div>
:::

Now, starting from that CSV (pretend someone emailed it to you):

a. Read `week4_cars.csv` (Python: `week4_cars_py.csv`) with `read_csv()` (Python:
   `pd.read_csv()`).
b. In one `mutate()` (Python: one `.assign()`): convert `am` to a factor (Python: categorical)
   labelled `"automatic"`/`"manual"`, `vs` to one labelled `"V-shaped"`/`"straight"`, and add
   `efficiency = mpg / hp`.
c. Produce a per-`cyl` summary table: number of cars, mean and standard deviation of `mpg`.
d. Draw a scatter plot of `hp` against `mpg`, coloured by transmission, faceted by `cyl`, with a
   least-squares line per transmission (no confidence band), full labels, and `theme_minimal()`.
e. Save the figure to `week4_analysis.png` (Python: `week4_analysis_py.png`) with `ggsave()`
   (Python: `.save()`).

Bonus points if (a)–(b) happen in a single pipeline.
:::

::: {.callout-important title="Challenge 4.2: Waves, functionally"}
A plotting challenge that resurrects @sec-week3. Reproduce this figure:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
#| echo: false
wave_x <- seq(0, 10, by = 0.01)
wave <- \(fr) bind_rows(
  tibble(x = wave_x, y = sin(fr * x), func = "sin", freq = fr),
  tibble(x = wave_x, y = cos(fr * x), func = "cos", freq = fr)
)
waves <- map(c(0.5, 1, 2), wave) |> bind_rows()
ggplot(waves, aes(x = x, y = y, colour = func)) +
  geom_line() +
  facet_wrap(~ freq, ncol = 1, labeller = label_both) +
  theme_minimal()
```

</div>

<div>

**Python**

```{python}
#| echo: false
wave_x = np.linspace(0, 10, 1001)
wave = lambda fr: pd.concat([
    pd.DataFrame({"x": wave_x, "y": np.sin(fr * wave_x), "func": "sin", "freq": fr}),
    pd.DataFrame({"x": wave_x, "y": np.cos(fr * wave_x), "func": "cos", "freq": fr})
])
waves = pd.concat([wave(fr) for fr in [0.5, 1, 2]])
(
  ggplot(waves, aes(x="x", y="y", colour="func")) +
  geom_line() +
  facet_wrap("~freq", ncol=1, labeller="label_both") +
  theme_minimal()
).show()
```

</div>
:::

That is six curves ($\sin(fx)$ and $\cos(fx)$ for frequencies $f \in \{0.5, 1, 2\}$ over
$x \in [0, 10]$) in one data frame with columns `x`, `y`, `func` and `freq`, plotted with
colour mapped to the function and one facet per frequency (`labeller = label_both` prints
`freq: 0.5` etc. in the panel strips).

The rules: no copy-pasting six blocks of data-frame code, and no `for` loops. Write a lambda
that builds the data frame for one frequency, `map()` it over the three frequencies (Python:
a list comprehension does the same job), and stitch the results with `bind_rows()` (Python:
`pd.concat()`), both of which happily accept a list of data frames.
:::

That completes the toolbox: data structures, control flow, functions, and now data wrangling
and graphics. Next week (@sec-week5) we put everything to work!
