# Week 5: Optimisation for Statistics {#sec-week5}

```{r}
#| include: false
set.seed(7432)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, out.width = "70%", fig.align = "center")
```

```{python}
#| include: false
import numpy as np
rng = np.random.default_rng(7432)
import matplotlib
matplotlib.rcParams["figure.figsize"] = (6, 4)
```

This is the week where everything pays off. In @sec-week2 we met gradient descent in
one dimension, in @sec-week3 we wrapped it into a reusable function, and in @sec-week4
we learnt to draw pictures of what our code is doing. This week we put those tools to
work on the problem that underlies an enormous amount of modern statistics:

> **Fitting a model is minimising a function.**

Maximum likelihood, least squares, penalised regression, neural networks — behind each
of them sits an optimisation problem, and behind most of the algorithms that solve
those problems sits some descendant of gradient descent. By the end of this chapter you
will have met the whole family: plain gradient descent, its stochastic cousin, momentum,
preconditioning, and the quasi-Newton method BFGS that the built-in optimisers of both
languages — R's `optim()` and Python's `scipy.optimize.minimize()` — use in practice.

## Gradient descent, again {#sec-week5-gd}

Quick recap. We want to minimise a differentiable function
$f : \mathbb{R}^d \to \mathbb{R}$. The gradient $\nabla f(\mathbf{x})$ points in the
direction of steepest *increase*, so we repeatedly take a small step the other way:

$$
\mathbf{x}_{n+1} = \mathbf{x}_n - \gamma \, \nabla f(\mathbf{x}_n),
$$

where $\gamma > 0$ is the **step size** (machine learning people say *learning rate*).
In @sec-week3 we packaged this into `grad_descent(f, grad, x0, gamma, n)`. Here it is
again — chapters are self-contained — with one upgrade: instead of returning only the
final point, it records the whole trajectory, because this week we want to *watch*
the algorithm work.

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
grad_descent <- function(f, grad, x0, gamma, n) {
  # one row per iterate, one column per coordinate
  path <- matrix(NA_real_, nrow = n + 1, ncol = length(x0))
  path[1, ] <- x0
  for (i in 1:n) {
    path[i + 1, ] <- path[i, ] - gamma * grad(path[i, ])
  }
  # return a data frame ready for ggplot: iterates plus the value of f at each
  out <- as.data.frame(path)
  names(out) <- paste0("x", seq_along(x0))
  out$f <- apply(path, 1, f)
  out$iter <- 0:n
  out
}
```

</div>

<div>

**Python**

```{python}
import numpy as np
import pandas as pd

def grad_descent(f, grad, x0, gamma, n):
    # one row per iterate, one column per coordinate
    path = np.full((n + 1, len(x0)), np.nan)
    path[0] = x0
    for i in range(n):
        path[i + 1] = path[i] - gamma * grad(path[i])
    # return a data frame ready for plotting: iterates plus the value of f at each
    out = pd.DataFrame(path, columns=[f"x{j + 1}" for j in range(len(x0))])
    out["f"] = [f(row) for row in path]
    out["iter"] = range(n + 1)
    return out
```

</div>
:::

**Note for Python users:** the R loop runs `i = 1, ..., n` and fills rows
`2, ..., n + 1`; the Python loop runs `i = 0, ..., n - 1` and fills rows `1, ..., n`
— the same trajectory, one index base apart (@sec-week1's indexing note at work).

Note the functional style from @sec-week3: `f` and `grad` are *arguments*. The same
dozen lines will minimise a toy parabola now and a logistic regression likelihood
later, without changing a character.

Let us try it on a friendly two-dimensional bowl,

$$
f(x_1, x_2) = (x_1 - 1)^2 + 2\,(x_2 + 1)^2,
\qquad
\nabla f(x_1, x_2) = \begin{pmatrix} 2(x_1 - 1) \\ 4(x_2 + 1) \end{pmatrix},
$$

whose minimum sits at $(1, -1)$.

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
f_bowl <- \(x) (x[1] - 1)^2 + 2 * (x[2] + 1)^2
grad_bowl <- \(x) c(2 * (x[1] - 1), 4 * (x[2] + 1))

path_bowl <- grad_descent(f_bowl, grad_bowl, x0 = c(-2, 2), gamma = 0.1, n = 60)
round(tail(path_bowl, 3), 5)
```

</div>

<div>

**Python**

```{python}
f_bowl = lambda x: (x[0] - 1)**2 + 2 * (x[1] + 1)**2
grad_bowl = lambda x: np.array([2 * (x[0] - 1), 4 * (x[1] + 1)])

path_bowl = grad_descent(f_bowl, grad_bowl, x0=np.array([-2, 2]), gamma=0.1, n=60)
path_bowl.tail(3).round(5)
```

</div>
:::

Sixty steps and we are sitting on the minimum. To *see* the descent we plot the
function as a **contour plot**: evaluate $f$ on a grid — `expand.grid()` in R,
`np.meshgrid()` in Python — then hand the grid to `geom_contour()`. Since we will
draw many of these, we write two small helper functions (again: @sec-week3 habits
paying rent).

**Note for Python users:** plotnine, ggplot2's Python twin from @sec-week4, has no
`geom_contour()`. Every contour figure this week therefore falls back on matplotlib's
`plt.contour()` — same pictures, different engine — and the faceted contour figures
become `plt.subplots()` panels.

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
library(ggplot2)

# evaluate f on a grid, ready for geom_contour
contour_data <- function(f, xlim, ylim, n = 100) {
  grid <- expand.grid(x1 = seq(xlim[1], xlim[2], length.out = n),
                      x2 = seq(ylim[1], ylim[2], length.out = n))
  grid$f <- apply(grid[, 1:2], 1, f)
  grid
}

# grey contours of f, with the trajectory drawn on top
plot_path <- function(f, path, xlim, ylim, bins = 20, colour = "#0072B2") {
  ggplot(contour_data(f, xlim, ylim), aes(x1, x2)) +
    geom_contour(aes(z = f), bins = bins, colour = "grey70") +
    geom_path(data = path, colour = colour, linewidth = 0.4) +
    geom_point(data = path, colour = colour, size = 0.9) +
    theme_minimal()
}

plot_path(f_bowl, path_bowl, xlim = c(-2.5, 3), ylim = c(-3, 2.5))
```

</div>

<div>

**Python**

```{python}
import matplotlib.pyplot as plt

# Unfortunately geom_contour does not work on plotnine!
# There are alternatives using matplotlib.

# evaluate f on a grid, ready for plt.contour
def contour_data(f, xlim, ylim, n=100):
    x1, x2 = np.meshgrid(np.linspace(xlim[0], xlim[1], n),
                         np.linspace(ylim[0], ylim[1], n))
    f_grid = np.array([f(np.array([a, b]))
                       for a, b in zip(x1.ravel(), x2.ravel())])
    return x1, x2, f_grid.reshape(x1.shape)

# grey contours of f, with the trajectory drawn on top
def plot_path(f, path, xlim, ylim, bins=20, colour="#0072B2"):
    x1, x2, f_grid = contour_data(f, xlim, ylim)
    fig, ax = plt.subplots()
    ax.contour(x1, x2, f_grid, levels=bins, colors="0.7", linewidths=0.5)
    ax.plot(path["x1"], path["x2"], color=colour, linewidth=0.8)
    ax.scatter(path["x1"], path["x2"], color=colour, s=8)
    ax.set_xlabel("x1")
    ax.set_ylabel("x2")
    return fig, ax

fig, ax = plot_path(f_bowl, path_bowl, xlim=(-2.5, 3), ylim=(-3, 2.5))
plt.show()
```

</div>
:::

The path marches straight into the bowl. Lovely. Time to break it.

## Why the step size is hard {#sec-week5-stepsize}

The step size $\gamma$ is doing a lot of quiet work. To expose it, we use a function
that is deliberately *ill-conditioned* — much steeper in one direction than the other:

$$
f(x_1, x_2) = \frac{x_1^2}{2} + \frac{10\, x_2^2}{2},
\qquad
\nabla f(x_1, x_2) = \begin{pmatrix} x_1 \\ 10\,x_2 \end{pmatrix}.
$$

Think of a long narrow valley: gentle along $x_1$ (curvature $1$), steep across it in
$x_2$ (curvature $10$). Watch what three different step sizes do, starting from
$(-9, 1.5)$.

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
#| fig-width: 9
#| fig-height: 3.4
#| out-width: "100%"
f_ill <- \(x) x[1]^2 / 2 + 10 * x[2]^2 / 2
grad_ill <- \(x) c(x[1], 10 * x[2])

path_small <- grad_descent(f_ill, grad_ill, c(-9, 1.5), gamma = 0.01, n = 60)
path_zig   <- grad_descent(f_ill, grad_ill, c(-9, 1.5), gamma = 0.18, n = 60)
path_boom  <- grad_descent(f_ill, grad_ill, c(-9, 1.5), gamma = 0.25, n = 8)

# one labelled data frame, one facet per step size (week 4 tricks)
path_small$gamma <- "gamma = 0.01: crawling"
path_zig$gamma   <- "gamma = 0.18: zig-zag"
path_boom$gamma  <- "gamma = 0.25: boom"
paths3 <- rbind(path_small, path_zig, path_boom)
paths3$gamma <- factor(paths3$gamma, levels = unique(paths3$gamma))

# the contour layer has no `gamma` column, so it is drawn in every facet
ggplot(contour_data(f_ill, c(-10, 3), c(-4, 4)), aes(x1, x2)) +
  geom_contour(aes(z = f), bins = 20, colour = "grey70") +
  geom_path(data = paths3, colour = "#0072B2", linewidth = 0.4) +
  geom_point(data = paths3, colour = "#0072B2", size = 0.9) +
  coord_cartesian(ylim = c(-4, 4)) +
  facet_wrap(~ gamma) +
  theme_minimal()
```

</div>

<div>

**Python**

```{python}
f_ill = lambda x: x[0]**2 / 2 + 10 * x[1]**2 / 2
grad_ill = lambda x: np.array([x[0], 10 * x[1]])

path_small = grad_descent(f_ill, grad_ill, np.array([-9, 1.5]), gamma=0.01, n=60)
path_zig   = grad_descent(f_ill, grad_ill, np.array([-9, 1.5]), gamma=0.18, n=60)
path_boom  = grad_descent(f_ill, grad_ill, np.array([-9, 1.5]), gamma=0.25, n=8)

# one list of labelled paths, one subplot per step size (no facets in matplotlib)
titles = ["gamma = 0.01: crawling", "gamma = 0.18: zig-zag", "gamma = 0.25: boom"]
paths3 = [path_small, path_zig, path_boom]

# the same contours are drawn in every panel, then one trajectory per panel
x1, x2, f_grid = contour_data(f_ill, (-10, 3), (-4, 4))
fig, axes = plt.subplots(1, 3, figsize=(9, 3.4), sharey=True)
for ax, p, ttl in zip(axes, paths3, titles):
    ax.contour(x1, x2, f_grid, levels=20, colors="0.7", linewidths=0.5)
    ax.plot(p["x1"], p["x2"], color="#0072B2", linewidth=0.8)
    ax.scatter(p["x1"], p["x2"], color="#0072B2", s=8)
    ax.set_ylim(-4, 4)
    ax.set_title(ttl, fontsize=9)
plt.show()
```

</div>
:::

Three step sizes, three personalities:

- **Too small** ($\gamma = 0.01$): every step is safe and almost useless. After 60
  iterations we have barely moved along the valley.
- **Near the limit** ($\gamma = 0.18$): the steep $x_2$ direction overshoots and
  bounces from one wall of the valley to the other — the famous **zig-zag** — while
  slowly drifting towards the minimum.
- **Too big** ($\gamma = 0.25$): each bounce is *bigger* than the last. The iterates
  diverge. (The panel is clipped; the true trajectory left the building.)

### The condition number {#sec-week5-conditioning}

There is a clean piece of intuition behind this. For a quadratic, gradient descent in
each coordinate multiplies the error by $(1 - \gamma \lambda)$ per step, where
$\lambda$ is the curvature in that direction. (These directional curvatures are
exactly the *eigenvalues* of the Hessian of $f$ — the `eigen()` from @sec-week1,
quietly reappearing as promised.) Stability requires
$|1 - \gamma \lambda| < 1$ for *every* direction, i.e.

$$
\gamma < \frac{2}{L}, \qquad L = \text{largest curvature},
$$

which for our valley means $\gamma < 2/10 = 0.2$ — exactly why $0.18$ scraped through
and $0.25$ exploded. But *progress* in the flattest direction is governed by the
smallest curvature $\mu$: per step the error shrinks by $(1 - \gamma\mu)$, at best
about $(1 - \mu/L)$. The ratio

$$
\kappa = \frac{L}{\mu} \qquad \text{(the condition number)}
$$

controls how many iterations you need: as $\kappa$ grows, the iteration count grows
in proportion. Our valley has $\kappa = 10$ and is already annoying. Real statistical problems —
say a regression with one covariate measured in grams and another in tonnes — can have
$\kappa$ in the millions, and plain gradient descent simply stops being a usable tool.
Hold that thought; it is the villain for the rest of the chapter.

A small helper we will use to compare methods fairly: how many iterations until we are
within `tol` of the target?

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
steps_to_tol <- function(path, target, tol = 1e-3) {
  X <- as.matrix(path[, grep("^x", names(path))])
  dist <- sqrt(colSums((t(X) - target)^2))
  hit <- which(dist < tol)
  if (length(hit) == 0) NA_integer_ else path$iter[hit[1]]
}

c(crawl = steps_to_tol(grad_descent(f_ill, grad_ill, c(-9, 1.5), 0.01, 5000), c(0, 0)),
  zig   = steps_to_tol(grad_descent(f_ill, grad_ill, c(-9, 1.5), 0.18, 5000), c(0, 0)))
```

</div>

<div>

**Python**

```{python}
def steps_to_tol(path, target, tol=1e-3):
    X = path.filter(regex="^x").to_numpy()
    dist = np.sqrt(((X - target)**2).sum(axis=1))
    hit = np.nonzero(dist < tol)[0]
    return np.nan if len(hit) == 0 else int(path["iter"].iloc[hit[0]])

{"crawl": steps_to_tol(grad_descent(f_ill, grad_ill, np.array([-9, 1.5]), 0.01, 5000),
                       np.array([0, 0])),
 "zig":   steps_to_tol(grad_descent(f_ill, grad_ill, np.array([-9, 1.5]), 0.18, 5000),
                       np.array([0, 0]))}
```

</div>
:::

::: {.callout-tip title="Exercise 5.1: Goldilocks and the step size"}
Consider $f(x_1, x_2) = x_1^2/2 + 25\,x_2^2/2$, with gradient
$\nabla f = (x_1, \, 25 x_2)$, starting point $(-8, 2)$, and true minimum $(0,0)$.

a. Using the stability condition $\gamma < 2/L$, predict the largest stable step
   size. Check your prediction by running `grad_descent()` with $\gamma$ just below
   and just above it (a small `n`, say 100, is enough to see divergence).
b. Use `steps_to_tol()` to count the iterations needed to get within $10^{-3}$ of
   the minimum for $\gamma = 0.005, 0.04$ and $0.07$ (use a generous `n`, e.g.
   20000). What do you conclude about how to choose $\gamma$?
:::

## Estimation is optimisation {#sec-week5-losses}

So far $f$ has been a toy. Here is why statisticians care. A huge share of estimators
are defined as the minimiser of an **average loss over the data**:

$$
\hat{\theta} = \arg\min_{\theta} L(\theta),
\qquad
L(\theta) = \frac{1}{n} \sum_{i=1}^{n} \ell_i(\theta),
$$

where $\ell_i(\theta)$ measures how badly the parameter $\theta$ explains observation
$i$. Two examples will follow us for the rest of the chapter.

**Linear regression (mean squared error).** With observations
$(\mathbf{x}_i, y_i)$ and model $y_i \approx \mathbf{x}_i^\top \boldsymbol\beta$, take
$\ell_i(\boldsymbol\beta) = \tfrac{1}{2}(y_i - \mathbf{x}_i^\top \boldsymbol\beta)^2$, so

$$
L(\boldsymbol\beta) = \frac{1}{2n} \|\mathbf{y} - X\boldsymbol\beta\|^2,
\qquad
\nabla L(\boldsymbol\beta) = -\frac{1}{n} X^\top (\mathbf{y} - X\boldsymbol\beta).
$$

(The $\tfrac12$ is cosmetic: it cancels the 2 from differentiating the square.
Minimising $L$ is least squares, which for Gaussian errors is also maximum likelihood.)

**Logistic regression (negative log-likelihood).** For binary $y_i \in \{0, 1\}$ the
model is $\Pr(y_i = 1 \mid \mathbf{x}_i) = \sigma(\mathbf{x}_i^\top\boldsymbol\beta)$
with $\sigma(z) = 1/(1 + e^{-z})$. Writing out $-\log$ of the likelihood and dividing
by $n$:

$$
L(\boldsymbol\beta) = \frac{1}{n}\sum_{i=1}^n
  \Big[ \log\!\big(1 + e^{\mathbf{x}_i^\top\boldsymbol\beta}\big)
        - y_i \, \mathbf{x}_i^\top\boldsymbol\beta \Big],
\qquad
\nabla L(\boldsymbol\beta) = \frac{1}{n} X^\top\big(\sigma(X\boldsymbol\beta) - \mathbf{y}\big),
$$

with $\sigma$ applied elementwise. No closed-form minimiser exists here — you *must*
optimise numerically. This is the everyday reality of maximum likelihood.

Let us simulate one dataset for each. We fix the random seed first — `set.seed()` in
R, a seeded generator `np.random.default_rng()` in Python — so that you get the same
data, plots and fitted values as these notes.

**Note:** the two languages cannot share a random stream. The R and Python columns
draw *different* datasets from the same model: each column is reproducible within its
own language, and every fitted value below matches its own column's data, but the
numbers differ slightly between the columns.

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
set.seed(7432)

# --- linear regression data: y = 2 - x + noise ------------------------------
n_obs <- 200
x_reg <- runif(n_obs, -2, 2)
X <- cbind(1, x_reg)                      # design matrix: intercept + slope
beta_true <- c(2, -1)
y <- as.vector(X %*% beta_true + rnorm(n_obs, sd = 0.8))

mse_loss <- \(b) sum((y - X %*% b)^2) / (2 * n_obs)
mse_grad <- \(b) as.vector(-t(X) %*% (y - X %*% b)) / n_obs

# --- logistic regression data: 2 covariates + intercept ---------------------
n_log <- 500
X_log <- cbind(1, matrix(rnorm(n_log * 2), ncol = 2))
beta_log_true <- c(-0.5, 2, -1)
p_true <- 1 / (1 + exp(-as.vector(X_log %*% beta_log_true)))
y_log <- rbinom(n_log, 1, p_true)

sigmoid <- \(z) 1 / (1 + exp(-z))
nll <- \(b) {
  eta <- as.vector(X_log %*% b)
  mean(log(1 + exp(eta)) - y_log * eta)
}
nll_grad <- \(b) {
  p <- sigmoid(as.vector(X_log %*% b))
  as.vector(t(X_log) %*% (p - y_log)) / n_log
}
```

</div>

<div>

**Python**

```{python}
rng = np.random.default_rng(7432)

# --- linear regression data: y = 2 - x + noise ------------------------------
n_obs = 200
x_reg = rng.uniform(-2, 2, n_obs)
X = np.column_stack([np.ones(n_obs), x_reg])   # design matrix: intercept + slope
beta_true = np.array([2, -1])
y = X @ beta_true + rng.normal(0, 0.8, n_obs)

mse_loss = lambda b: np.sum((y - X @ b)**2) / (2 * n_obs)
mse_grad = lambda b: -X.T @ (y - X @ b) / n_obs

# --- logistic regression data: 2 covariates + intercept ---------------------
n_log = 500
X_log = np.column_stack([np.ones(n_log), rng.normal(size=(n_log, 2))])
beta_log_true = np.array([-0.5, 2, -1])
p_true = 1 / (1 + np.exp(-(X_log @ beta_log_true)))
y_log = rng.binomial(1, p_true)

sigmoid = lambda z: 1 / (1 + np.exp(-z))
def nll(b):
    eta = X_log @ b
    return np.mean(np.log(1 + np.exp(eta)) - y_log * eta)

def nll_grad(b):
    p = sigmoid(X_log @ b)
    return X_log.T @ (p - y_log) / n_log
```

</div>
:::

**Note for Python users:** a `lambda` holds a single expression only, so the
multi-line R lambdas become small `def`s — the same first-class function objects
either way (@sec-week3).

For linear regression we do not actually *need* an iterative method: setting
$\nabla L = 0$ gives the normal equations
$X^\top X \boldsymbol\beta = X^\top \mathbf{y}$, which `solve()` — `np.linalg.solve()`
in Python — dispatches in one line, the very linear algebra you did in @sec-week1:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
beta_ols <- as.vector(solve(t(X) %*% X, t(X) %*% y))
beta_ols
```

</div>

<div>

**Python**

```{python}
beta_ols = np.linalg.solve(X.T @ X, X.T @ y)
beta_ols
```

</div>
:::

(In real life R's `lm()` — or `np.linalg.lstsq()` in Python — does this for you, more
carefully.) The closed form gives us a gold standard to check gradient descent against:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
path_mse <- grad_descent(mse_loss, mse_grad, x0 = c(0, 0), gamma = 0.5, n = 100)

rbind(closed_form = beta_ols,
      grad_descent = unlist(tail(path_mse, 1)[1:2]))
```

</div>

<div>

**Python**

```{python}
path_mse = grad_descent(mse_loss, mse_grad, x0=np.array([0.0, 0.0]), gamma=0.5, n=100)

pd.DataFrame([beta_ols, path_mse.iloc[-1, :2].to_numpy()],
             index=["closed_form", "grad_descent"], columns=["x1", "x2"])
```

</div>
:::

Identical to working precision, in about fifty effective steps. And because the loss
has two parameters, we can watch the descent over the loss surface itself — the
contours below are contours of $L(\beta_0, \beta_1)$, a picture worth keeping in mind
every time you fit *any* model:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
plot_path(mse_loss, path_mse, xlim = c(-0.5, 3), ylim = c(-2.5, 0.7)) +
  geom_point(data = data.frame(x1 = beta_ols[1], x2 = beta_ols[2]),
             shape = 4, size = 3, colour = "black") +
  labs(x = expression(beta[0]), y = expression(beta[1]))
```

</div>

<div>

**Python**

```{python}
fig, ax = plot_path(mse_loss, path_mse, xlim=(-0.5, 3), ylim=(-2.5, 0.7))
ax.scatter([beta_ols[0]], [beta_ols[1]], marker="x", s=60, color="black")
ax.set_xlabel(r"$\beta_0$")
ax.set_ylabel(r"$\beta_1$")
plt.show()
```

</div>
:::

The cross marks $\hat{\boldsymbol\beta}_{\text{OLS}}$. Gradient descent walks straight
to the same place the algebra pointed to. For logistic regression there is no cross to
draw in advance — the iterative method is the only game in town. We will fit it in
@sec-week5-bfgs.

## Stochastic gradient descent {#sec-week5-sgd}

Look again at the gradient of the average loss:

$$
\nabla L(\theta) = \frac{1}{n} \sum_{i=1}^{n} \nabla \ell_i(\theta).
$$

Every single step of gradient descent touches **all $n$ observations**. For our
$n = 200$ toy that is nothing; for $n = 10^8$ user records it means one step of GD
costs a full pass over a dataset that may not even fit in memory. The sum structure
suggests a cheeky fix: *don't compute the whole sum*. Pick one index $I$ uniformly at
random and use $\nabla \ell_I(\theta)$ as a stand-in for $\nabla L(\theta)$. It is a
noisy stand-in, but it is **unbiased**:
$\mathbb{E}\big[\nabla \ell_I(\theta)\big] = \nabla L(\theta)$ — on average, it is the
right direction. The update

$$
\theta_{t+1} = \theta_t - \gamma_t \, \nabla \ell_{I_t}(\theta_t),
\qquad I_t \sim \text{Uniform}\{1, \dots, n\},
$$

is **stochastic gradient descent** (SGD). A compromise between one sample and all $n$
is a **mini-batch**: average $\nabla\ell_i$ over a random subset $B$ of size $m$. Still
unbiased, with variance shrinking like $1/m$, at cost proportional to $m$. The
economics are the whole point:

- one GD step costs $O(n)$ per-observation gradients;
- one SGD step costs $O(1)$ (or $O(m)$ for a mini-batch), *regardless of $n$*.

When $n$ is huge, SGD has taken thousands of noisy-but-useful steps before GD finishes
its first.

Here is single-sample SGD in the same functional style. One new ingredient: the step
size is now a *function* of the step number $t$, for reasons one paragraph away.

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
sgd <- function(grad_i, x0, gamma, n_steps, n_data) {
  # grad_i(theta, i): gradient of the loss at observation i only
  # gamma(t): step size at step t
  path <- matrix(NA_real_, nrow = n_steps + 1, ncol = length(x0))
  path[1, ] <- x0
  for (t in 1:n_steps) {
    i <- sample(n_data, 1)
    path[t + 1, ] <- path[t, ] - gamma(t) * grad_i(path[t, ], i)
  }
  out <- as.data.frame(path)
  names(out) <- paste0("x", seq_along(x0))
  out$iter <- 0:n_steps
  out
}

# single-observation gradient for the regression loss
mse_grad_i <- function(b, i) as.vector(-X[i, ] * (y[i] - sum(X[i, ] * b)))
```

</div>

<div>

**Python**

```{python}
def sgd(grad_i, x0, gamma, n_steps, n_data):
    # grad_i(theta, i): gradient of the loss at observation i only
    # gamma(t): step size at step t
    path = np.full((n_steps + 1, len(x0)), np.nan)
    path[0] = x0
    for t in range(1, n_steps + 1):
        i = rng.integers(n_data)
        path[t] = path[t - 1] - gamma(t) * grad_i(path[t - 1], i)
    out = pd.DataFrame(path, columns=[f"x{j + 1}" for j in range(len(x0))])
    out["iter"] = range(n_steps + 1)
    return out

# single-observation gradient for the regression loss
mse_grad_i = lambda b, i: -X[i] * (y[i] - X[i] @ b)
```

</div>
:::

**Note for Python users:** `sample(n_data, 1)` draws from $\{1, \dots, n\}$;
`rng.integers(n_data)` draws from $\{0, \dots, n-1\}$ — each exactly the right range
for its language's indexing. And where R's `sample()` quietly uses the global RNG
state, the Python column leans on the global `rng` object in the same way.

Let us run it on the regression loss twice: once with a constant step size, once with
a decaying one.

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
#| fig-width: 8
#| fig-height: 3.8
#| out-width: "100%"
set.seed(1)
sgd_const <- sgd(mse_grad_i, c(0, 0), gamma = \(t) 0.1,          n_steps = 2000, n_data = n_obs)
sgd_decay <- sgd(mse_grad_i, c(0, 0), gamma = \(t) 2 / (20 + t), n_steps = 2000, n_data = n_obs)

sgd_const$schedule <- "constant: gamma_t = 0.1"
sgd_decay$schedule <- "decaying: gamma_t = 2/(20 + t)"
sgd_both <- rbind(sgd_const, sgd_decay)

ols_pt <- data.frame(x1 = beta_ols[1], x2 = beta_ols[2])

ggplot(contour_data(mse_loss, c(-0.5, 3), c(-2.5, 0.7)), aes(x1, x2)) +
  geom_contour(aes(z = f), bins = 20, colour = "grey70") +
  geom_path(data = sgd_both, colour = "#009E73", linewidth = 0.4) +
  geom_point(data = ols_pt, shape = 4, size = 3, colour = "black") +
  facet_wrap(~ schedule) +
  labs(x = expression(beta[0]), y = expression(beta[1])) +
  theme_minimal()
```

</div>

<div>

**Python**

```{python}
rng = np.random.default_rng(1)
sgd_const = sgd(mse_grad_i, np.array([0.0, 0.0]), gamma=lambda t: 0.1,          n_steps=2000, n_data=n_obs)
sgd_decay = sgd(mse_grad_i, np.array([0.0, 0.0]), gamma=lambda t: 2 / (20 + t), n_steps=2000, n_data=n_obs)

# one subplot per schedule, same contours in each
titles = ["constant: gamma_t = 0.1", "decaying: gamma_t = 2/(20 + t)"]
x1, x2, f_grid = contour_data(mse_loss, (-0.5, 3), (-2.5, 0.7))
fig, axes = plt.subplots(1, 2, figsize=(8, 3.8), sharey=True)
for ax, p, ttl in zip(axes, [sgd_const, sgd_decay], titles):
    ax.contour(x1, x2, f_grid, levels=20, colors="0.7", linewidths=0.5)
    ax.plot(p["x1"], p["x2"], color="#009E73", linewidth=0.8)
    ax.scatter([beta_ols[0]], [beta_ols[1]], marker="x", s=60, color="black")
    ax.set_title(ttl, fontsize=9)
    ax.set_xlabel(r"$\beta_0$")
axes[0].set_ylabel(r"$\beta_1$")
plt.show()
```

</div>
:::

Both trajectories are drunk versions of the gradient descent path — that is the noise
of estimating the gradient from one observation. But the endgames differ:

- With a **constant** step the iterates never settle: they reach the neighbourhood of
  $\hat{\boldsymbol\beta}$ quickly and then rattle around it forever, in a "noise
  ball" whose radius grows with $\gamma$.
- With a **decaying** step the early iterations still move fast, while the shrinking
  steps gradually strangle the noise, and the trajectory funnels into the minimum.

How fast should $\gamma_t$ decay? The classical answer (Robbins and Monro, 1951 —
this algorithm is older than the computer you are running it on) is, informally:

$$
\sum_{t=1}^{\infty} \gamma_t = \infty
\qquad \text{and} \qquad
\sum_{t=1}^{\infty} \gamma_t^2 < \infty.
$$

The first condition says the steps must add up to infinity — otherwise the algorithm
can run out of fuel before reaching the minimum, however far away it started. The
second says they must shrink fast enough that the noise averages out. Schedules like
$\gamma_t = a/(b + t)$ satisfy both, which is why you keep seeing that shape.

Finally, the claim that actually sells SGD: progress *per data point touched*. One GD
step costs $n = 200$ observation-gradients; one SGD step costs 1. Plot the loss
against that budget:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
#| message: false
library(dplyr)

sgd_decay$loss <- apply(as.matrix(sgd_decay[, 1:2]), 1, mse_loss)

cost_df <- bind_rows(
  tibble(evals = path_mse$iter * n_obs, loss = path_mse$f,   method = "GD (full gradient)"),
  tibble(evals = sgd_decay$iter,        loss = sgd_decay$loss, method = "SGD (one sample)")
) |>
  filter(evals > 0)

ggplot(cost_df, aes(evals, loss, colour = method)) +
  geom_line(linewidth = 0.7) +
  scale_x_log10() +
  scale_colour_manual(values = c("GD (full gradient)" = "#0072B2",
                                 "SGD (one sample)" = "#009E73")) +
  labs(x = "observation-gradients computed (log scale)", y = "loss L") +
  theme_minimal()
```

</div>

<div>

**Python**

```{python}
from plotnine import (ggplot, aes, geom_line, scale_x_log10,
                      scale_colour_manual, labs, theme_minimal)

sgd_decay["loss"] = [mse_loss(b) for b in sgd_decay[["x1", "x2"]].to_numpy()]

cost_df = pd.concat([
    pd.DataFrame({"evals": path_mse["iter"] * n_obs, "loss": path_mse["f"],
                  "method": "GD (full gradient)"}),
    pd.DataFrame({"evals": sgd_decay["iter"], "loss": sgd_decay["loss"],
                  "method": "SGD (one sample)"})
])
cost_df = cost_df[cost_df["evals"] > 0]

(ggplot(cost_df, aes("evals", "loss", colour="method"))
 + geom_line(size=0.7)
 + scale_x_log10()
 + scale_colour_manual(values={"GD (full gradient)": "#0072B2",
                               "SGD (one sample)": "#009E73"})
 + labs(x="observation-gradients computed (log scale)", y="loss L")
 + theme_minimal()).show()
```

</div>
:::

**Note for Python users:** pandas has no `bind_rows()` or `filter()` verbs —
`pd.concat()` plus a boolean mask do the same job (@sec-week4) — and plotnine still
calls the line-width aesthetic `size`, which ggplot2 has since renamed `linewidth`.

By the time GD has finished its *first* step (200 evaluations), SGD has already
collapsed most of the loss. Read the other end honestly too: GD then converges to the
exact minimiser, while SGD closes in more slowly and less precisely. For small $n$ and
high accuracy, GD (or better, @sec-week5-bfgs) wins; for enormous $n$ and
statistical-scale accuracy — remember the estimate itself has sampling error of order
$1/\sqrt{n}$, so chasing the tenth decimal of the minimiser is vanity — SGD is the
only affordable option. It is the workhorse that trains essentially every neural
network.

:::: {.callout-tip title="Exercise 5.2: Mini-batch SGD"}
Fill in the blanks to turn `sgd()` into mini-batch SGD, then run it on the regression
loss. The batch gradient function is given — note (on the R side) the `drop = FALSE`:
when the batch holds a single index, `X[idx, ]` would silently collapse the one-row
matrix to a plain vector, and `drop = FALSE` keeps the matrix shape so the algebra
survives `batch = 1`:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
#| eval: false
mse_grad_batch <- function(b, idx) {
  Xb <- X[idx, , drop = FALSE]
  as.vector(-t(Xb) %*% (y[idx] - Xb %*% b)) / length(idx)
}

sgd_batch <- function(grad_batch, x0, gamma, n_steps, n_data, batch = 10) {
  path <- matrix(NA_real_, nrow = n_steps + 1, ncol = length(x0))
  path[1, ] <- x0
  for (t in 1:n_steps) {
    idx <- sample(____, ____)          # draw a mini-batch of indices
    path[t + 1, ] <- ____              # the usual update, batch gradient
  }
  out <- as.data.frame(path)
  names(out) <- paste0("x", seq_along(x0))
  out$iter <- 0:n_steps
  out
}
```

</div>

<div>

**Python**

```{python}
#| eval: false
def mse_grad_batch(b, idx):
    Xb = X[idx]
    return -Xb.T @ (y[idx] - Xb @ b) / len(idx)

def sgd_batch(grad_batch, x0, gamma, n_steps, n_data, batch=10):
    path = np.full((n_steps + 1, len(x0)), np.nan)
    path[0] = x0
    for t in range(1, n_steps + 1):
        idx = rng.choice(____, ____, replace=False)   # draw a mini-batch of indices
        path[t] = ____                                # the usual update, batch gradient
    out = pd.DataFrame(path, columns=[f"x{j + 1}" for j in range(len(x0))])
    out["iter"] = range(n_steps + 1)
    return out
```

</div>
:::

**Note for Python users:** numpy dodges the `drop = FALSE` trap — indexing with an
index *array* (`X[idx]`) always keeps the 2-D shape; only a bare scalar index drops
the dimension. The trap on your side is `rng.choice()`, which samples **with**
replacement unless told `replace=False`, whereas R's `sample()` is without
replacement by default.

Run 2000 steps with constant `gamma = \(t) 0.1` (`lambda t: 0.1`) for batch sizes 1,
10 and 50, and compare the wobble of the final 500 iterates (e.g. via `sd()` of each
coordinate — `.std()` in Python — or just by plotting the paths). What does the batch
size buy you, and what does it cost?
::::

## Momentum: a ball rolling downhill {#sec-week5-momentum}

Back to the zig-zag valley of @sec-week5-stepsize. The step size was capped by the
steep direction, progress was throttled along the flat one, and the trajectory wasted
its energy bouncing between walls. Physics suggests a fix. Imagine the iterate is not
a point teleporting around, but a **heavy ball rolling on the surface of $f$, with
friction**. A ball entering a narrow valley rattles briefly between the walls, the
rattling cancels itself out, and the ball settles into rolling *along* the valley —
gathering speed in any direction the slope keeps pushing.

To simulate that, give the iterate a velocity $\mathbf{v}$ that accumulates gradients
instead of forgetting them:

$$
\mathbf{v}_{t+1} = \beta\, \mathbf{v}_t - \gamma\, \nabla f(\mathbf{x}_t),
\qquad
\mathbf{x}_{t+1} = \mathbf{x}_t + \mathbf{v}_{t+1},
$$

with $\beta \in [0, 1)$. This is the **heavy-ball method**, or gradient descent with
**momentum**. The coefficient $\beta$ is the memory of the ball: each step the old
velocity is kept, scaled by $\beta$, so $1 - \beta$ acts as friction. $\beta = 0$ is
plain gradient descent (total friction — all velocity lost each step); $\beta \to 1$ is a frictionless
ball that overshoots and oscillates forever. In between, something useful happens:

- in the **zig-zag direction** the gradient flips sign every step, so successive
  contributions to $\mathbf{v}$ *cancel* — the oscillation damps itself;
- in the **consistent downhill direction** the contributions all point the same way
  and *accumulate* — the ball accelerates along the valley floor.

In code, two lines change:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
grad_descent_mom <- function(f, grad, x0, gamma, n, beta = 0.9) {
  path <- matrix(NA_real_, nrow = n + 1, ncol = length(x0))
  path[1, ] <- x0
  v <- numeric(length(x0))            # start at rest
  for (i in 1:n) {
    v <- beta * v - gamma * grad(path[i, ])
    path[i + 1, ] <- path[i, ] + v
  }
  out <- as.data.frame(path)
  names(out) <- paste0("x", seq_along(x0))
  out$f <- apply(path, 1, f)
  out$iter <- 0:n
  out
}
```

</div>

<div>

**Python**

```{python}
def grad_descent_mom(f, grad, x0, gamma, n, beta=0.9):
    path = np.full((n + 1, len(x0)), np.nan)
    path[0] = x0
    v = np.zeros(len(x0))               # start at rest
    for i in range(n):
        v = beta * v - gamma * grad(path[i])
        path[i + 1] = path[i] + v
    out = pd.DataFrame(path, columns=[f"x{j + 1}" for j in range(len(x0))])
    out["f"] = [f(row) for row in path]
    out["iter"] = range(n + 1)
    return out
```

</div>
:::

Same valley, same aggressive $\gamma = 0.18$, with and without momentum:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
path_plain <- grad_descent(f_ill, grad_ill, c(-9, 1.5), gamma = 0.18, n = 60)
path_mom   <- grad_descent_mom(f_ill, grad_ill, c(-9, 1.5), gamma = 0.18, n = 60, beta = 0.5)

both <- bind_rows(mutate(path_plain, method = "gradient descent"),
                  mutate(path_mom,   method = "with momentum"))

ggplot(contour_data(f_ill, c(-10, 3), c(-4, 4)), aes(x1, x2)) +
  geom_contour(aes(z = f), bins = 20, colour = "grey70") +
  geom_path(data = both, aes(colour = method), linewidth = 0.4) +
  geom_point(data = both, aes(colour = method), size = 0.9) +
  scale_colour_manual(values = c("gradient descent" = "#0072B2",
                                 "with momentum" = "#D55E00")) +
  theme_minimal()
```

</div>

<div>

**Python**

```{python}
path_plain = grad_descent(f_ill, grad_ill, np.array([-9, 1.5]), gamma=0.18, n=60)
path_mom   = grad_descent_mom(f_ill, grad_ill, np.array([-9, 1.5]), gamma=0.18, n=60, beta=0.5)

both = [(path_plain, "#0072B2", "gradient descent"),
        (path_mom,   "#D55E00", "with momentum")]

x1, x2, f_grid = contour_data(f_ill, (-10, 3), (-4, 4))
fig, ax = plt.subplots()
ax.contour(x1, x2, f_grid, levels=20, colors="0.7", linewidths=0.5)
for p, colour, label in both:
    ax.plot(p["x1"], p["x2"], color=colour, linewidth=0.8, label=label)
    ax.scatter(p["x1"], p["x2"], color=colour, s=8)
ax.legend()
plt.show()
```

</div>
:::

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
c(plain    = steps_to_tol(grad_descent(f_ill, grad_ill, c(-9, 1.5), 0.18, 3000), c(0, 0)),
  momentum = steps_to_tol(grad_descent_mom(f_ill, grad_ill, c(-9, 1.5), 0.18, 3000, beta = 0.5),
                          c(0, 0)))
```

</div>

<div>

**Python**

```{python}
{"plain":    steps_to_tol(grad_descent(f_ill, grad_ill, np.array([-9, 1.5]), 0.18, 3000),
                          np.array([0, 0])),
 "momentum": steps_to_tol(grad_descent_mom(f_ill, grad_ill, np.array([-9, 1.5]), 0.18, 3000,
                                           beta=0.5),
                          np.array([0, 0]))}
```

</div>
:::

The orange path shakes off the zig-zag within a few bounces and cruises down the
valley, reaching tolerance in half the iterations — and the gap widens as $\kappa$
grows. In practice $\beta$ is a tuning knob like $\gamma$: values around $0.9$ are
the folklore default in machine learning, but on a nearly-quadratic problem like this
a smaller $\beta$ damps faster (try $0.9$ here and watch it overshoot the minimum and
loop back).

### Stochastic momentum

Momentum drops straight into SGD — replace the full gradient with the single-sample
one:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
sgd_mom <- function(grad_i, x0, gamma, n_steps, n_data, beta = 0.9) {
  path <- matrix(NA_real_, nrow = n_steps + 1, ncol = length(x0))
  path[1, ] <- x0
  v <- numeric(length(x0))
  for (t in 1:n_steps) {
    i <- sample(n_data, 1)
    v <- beta * v - gamma(t) * grad_i(path[t, ], i)
    path[t + 1, ] <- path[t, ] + v
  }
  out <- as.data.frame(path)
  names(out) <- paste0("x", seq_along(x0))
  out$iter <- 0:n_steps
  out
}
```

</div>

<div>

**Python**

```{python}
def sgd_mom(grad_i, x0, gamma, n_steps, n_data, beta=0.9):
    path = np.full((n_steps + 1, len(x0)), np.nan)
    path[0] = x0
    v = np.zeros(len(x0))
    for t in range(1, n_steps + 1):
        i = rng.integers(n_data)
        v = beta * v - gamma(t) * grad_i(path[t - 1], i)
        path[t] = path[t - 1] + v
    out = pd.DataFrame(path, columns=[f"x{j + 1}" for j in range(len(x0))])
    out["iter"] = range(n_steps + 1)
    return out
```

</div>
:::

Now $\mathbf{v}$ is a running average of *noisy* gradients, so momentum also acts as a
noise filter: individual samples may point sideways, but their average points
downhill. One honest subtlety: with memory $\beta$, the steps effectively add up to
$\gamma/(1-\beta)$ worth of gradient — with $\beta = 0.9$ that is a factor of 10 — so
shrink $\gamma$ accordingly when you switch momentum on.

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
set.seed(2)
s_plain <- sgd(mse_grad_i, c(0, 0), gamma = \(t) 2 / (20 + t), n_steps = 2000, n_data = n_obs)
set.seed(2)
s_mom <- sgd_mom(mse_grad_i, c(0, 0), gamma = \(t) 0.2 / (20 + t), n_steps = 2000,
                 n_data = n_obs, beta = 0.9)

both_sgd <- bind_rows(mutate(s_plain, method = "SGD"),
                      mutate(s_mom,   method = "SGD + momentum"))

ggplot(contour_data(mse_loss, c(-0.5, 3), c(-2.5, 0.7)), aes(x1, x2)) +
  geom_contour(aes(z = f), bins = 20, colour = "grey70") +
  geom_path(data = both_sgd, aes(colour = method), linewidth = 0.4) +
  geom_point(data = ols_pt, shape = 4, size = 3, colour = "black") +
  scale_colour_manual(values = c("SGD" = "#009E73",
                                 "SGD + momentum" = "#D55E00")) +
  labs(x = expression(beta[0]), y = expression(beta[1])) +
  theme_minimal()
```

</div>

<div>

**Python**

```{python}
rng = np.random.default_rng(2)
s_plain = sgd(mse_grad_i, np.array([0.0, 0.0]), gamma=lambda t: 2 / (20 + t),
              n_steps=2000, n_data=n_obs)
rng = np.random.default_rng(2)
s_mom = sgd_mom(mse_grad_i, np.array([0.0, 0.0]), gamma=lambda t: 0.2 / (20 + t),
                n_steps=2000, n_data=n_obs, beta=0.9)

both_sgd = [(s_plain, "#009E73", "SGD"),
            (s_mom,   "#D55E00", "SGD + momentum")]

x1, x2, f_grid = contour_data(mse_loss, (-0.5, 3), (-2.5, 0.7))
fig, ax = plt.subplots()
ax.contour(x1, x2, f_grid, levels=20, colors="0.7", linewidths=0.5)
for p, colour, label in both_sgd:
    ax.plot(p["x1"], p["x2"], color=colour, linewidth=0.8, label=label)
ax.scatter([beta_ols[0]], [beta_ols[1]], marker="x", s=60, color="black")
ax.set_xlabel(r"$\beta_0$")
ax.set_ylabel(r"$\beta_1$")
ax.legend()
plt.show()
```

</div>
:::

Both end up at the same place, but the momentum path is visibly smoother — it turns
gently where plain SGD staggers. SGD-plus-momentum, with mini-batches, is more or less
*the* algorithm that trains deep neural networks.

::: {.callout-tip title="Exercise 5.3: Momentum pays off"}
Take Exercise 5.1's valley, $f(x_1, x_2) = x_1^2/2 + 25 x_2^2/2$ with
$\nabla f = (x_1,\, 25 x_2)$, start $(-8, 2)$, and fix $\gamma = 0.07$.

a. Run `grad_descent()` and `grad_descent_mom()` with `beta = 0.6` (use `n = 20000`)
   and report `steps_to_tol()` to within $10^{-3}$ of the minimum for each.
b. Plot the two trajectories on the contours of $f$ (adapting the momentum-comparison
   code above). Where in the picture is momentum earning its keep?
:::

## Preconditioning: fixing the geometry {#sec-week5-precond}

Momentum treats the symptom of ill-conditioning. Preconditioning treats the cause:
the coordinates live on the wrong scales. If the valley is ten times steeper in $x_2$
than in $x_1$, then take steps ten times smaller in $x_2$. Formally, pick a symmetric
positive-definite matrix $P$ and update

$$
\mathbf{x}_{t+1} = \mathbf{x}_t - \gamma \, P \, \nabla f(\mathbf{x}_t).
$$

$P$ reshapes the descent direction, stretching it in flat directions and shrinking it
in steep ones. Plain gradient descent is $P = I$.

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
grad_descent_pre <- function(f, grad, x0, gamma, n, P) {
  path <- matrix(NA_real_, nrow = n + 1, ncol = length(x0))
  path[1, ] <- x0
  for (i in 1:n) {
    path[i + 1, ] <- path[i, ] - gamma * as.vector(P %*% grad(path[i, ]))
  }
  out <- as.data.frame(path)
  names(out) <- paste0("x", seq_along(x0))
  out$f <- apply(path, 1, f)
  out$iter <- 0:n
  out
}
```

</div>

<div>

**Python**

```{python}
def grad_descent_pre(f, grad, x0, gamma, n, P):
    path = np.full((n + 1, len(x0)), np.nan)
    path[0] = x0
    for i in range(n):
        path[i + 1] = path[i] - gamma * (P @ grad(path[i]))
    out = pd.DataFrame(path, columns=[f"x{j + 1}" for j in range(len(x0))])
    out["f"] = [f(row) for row in path]
    out["iter"] = range(n + 1)
    return out
```

</div>
:::

What is the *ideal* $P$? Our valley has Hessian (matrix of second derivatives)
$H = \mathrm{diag}(1, 10)$: curvature 1 along $x_1$, 10 along $x_2$. Choosing
$P = H^{-1} = \mathrm{diag}(1, 1/10)$ divides each direction by its own curvature, so
*every* direction behaves like curvature 1 — the valley becomes a perfectly round
bowl, $\kappa = 1$. With $\gamma = 1$:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
P <- diag(c(1, 1 / 10))
path_pre <- grad_descent_pre(f_ill, grad_ill, c(-9, 1.5), gamma = 1, n = 5, P = P)
round(path_pre, 10)
```

</div>

<div>

**Python**

```{python}
P = np.diag([1, 1 / 10])
path_pre = grad_descent_pre(f_ill, grad_ill, np.array([-9, 1.5]), gamma=1, n=5, P=P)
path_pre.round(10)
```

</div>
:::

One step. Not "converges quickly" — *lands exactly on the minimum in one step*.

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
both_pre <- bind_rows(mutate(path_zig, method = "GD, gamma = 0.18"),
                      mutate(path_pre, method = "preconditioned, gamma = 1"))

ggplot(contour_data(f_ill, c(-10, 3), c(-4, 4)), aes(x1, x2)) +
  geom_contour(aes(z = f), bins = 20, colour = "grey70") +
  geom_path(data = both_pre, aes(colour = method), linewidth = 0.4) +
  geom_point(data = both_pre, aes(colour = method), size = 0.9) +
  scale_colour_manual(values = c("GD, gamma = 0.18" = "#0072B2",
                                 "preconditioned, gamma = 1" = "#009E73")) +
  theme_minimal()
```

</div>

<div>

**Python**

```{python}
both_pre = [(path_zig, "#0072B2", "GD, gamma = 0.18"),
            (path_pre, "#009E73", "preconditioned, gamma = 1")]

x1, x2, f_grid = contour_data(f_ill, (-10, 3), (-4, 4))
fig, ax = plt.subplots()
ax.contour(x1, x2, f_grid, levels=20, colors="0.7", linewidths=0.5)
for p, colour, label in both_pre:
    ax.plot(p["x1"], p["x2"], color=colour, linewidth=0.8, label=label)
    ax.scatter(p["x1"], p["x2"], color=colour, s=8)
ax.legend()
plt.show()
```

</div>
:::

Taking $P = (\nabla^2 f)^{-1}$, recomputed at every iterate, with $\gamma = 1$, has a
name: **Newton's method**. On a quadratic it finishes in one step; near the minimum of
a smooth function it converges *quadratically* (the number of correct digits roughly
doubles per iteration). So why isn't Newton the end of the chapter?

- **Cost.** In $d$ dimensions the Hessian has $d^2$ entries, and applying its inverse
  means solving a $d \times d$ linear system: $O(d^3)$ per step. For a neural network
  with $d = 10^6$ parameters, the Hessian alone would need terabytes of memory.
- **Fragility.** Away from the minimum the Hessian need not be positive definite, and
  then $-H^{-1}\nabla f$ may not even point downhill. Making Newton robust takes real
  engineering.

Cheap compromises are everywhere in statistics and machine learning: use only the
*diagonal* of the Hessian (exactly our $P$ above — for axis-aligned problems it is the
whole story, and in general it is a decent approximation); or simply **standardise
your covariates**, which is preconditioning by a diagonal matrix before the algorithm
even starts. If you remember one practical trick from this section: *rescale your
variables before optimising*. Exercise 5.4 shows what happens if you don't.

## BFGS: learning the Hessian as you go {#sec-week5-bfgs}

Newton's method needs second derivatives we cannot afford. But here is the clever
observation: the gradients we are *already computing* contain second-order
information. By Taylor expansion, for nearby iterates,

$$
\nabla f(\mathbf{x}_{k+1}) - \nabla f(\mathbf{x}_k)
\;\approx\;
H \, (\mathbf{x}_{k+1} - \mathbf{x}_k),
$$

so each step tells us how the gradient *changed* when we moved in one particular
direction — one slice of the Hessian, for free. Quasi-Newton methods maintain a
running approximation $B_k \approx H^{-1}$ and require it to explain the most recent
step exactly, the **secant condition**:

$$
B_{k+1} \, \underbrace{\big(\nabla f_{k+1} - \nabla f_k\big)}_{\mathbf{y}_k}
= \underbrace{\mathbf{x}_{k+1} - \mathbf{x}_k}_{\mathbf{s}_k}.
$$

**BFGS** (Broyden–Fletcher–Goldfarb–Shanno, 1970 — four people found it independently
in the same year) updates $B_k$ by the smallest symmetric modification that satisfies
the secant condition — a *rank-two update*: cheap matrix–vector algebra, no matrix
inversion, and $B_k$ stays positive definite so the step $-B_k \nabla f_k$ always
points downhill (provided each step actually observes positive curvature,
$\mathbf{y}_k^\top \mathbf{s}_k > 0$ — the line search inside `optim()` and
`minimize()` sees to this). For the curious, the update is

$$
B_{k+1} = \big(I - \rho_k \mathbf{s}_k \mathbf{y}_k^\top\big) B_k
          \big(I - \rho_k \mathbf{y}_k \mathbf{s}_k^\top\big)
          + \rho_k \mathbf{s}_k \mathbf{s}_k^\top,
\qquad \rho_k = \frac{1}{\mathbf{y}_k^\top \mathbf{s}_k},
$$

and we will not derive it — the point to retain is the *idea*: start with $B_0 = I$
(plain gradient descent), and let each observed gradient difference teach the
algorithm a bit more of the local geometry. After a handful of steps BFGS is taking
nearly Newton-quality steps while having only ever evaluated $f$ and $\nabla f$.

You will never implement BFGS yourself, because both languages ship it — R as
`optim()`, Python inside `scipy.optimize.minimize()`:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```r
optim(par, fn, gr, method = "BFGS")
```

</div>

<div>

**Python**

```python
sp.optimize.minimize(fun, x0, jac=grad, method="BFGS")
```

</div>
:::

with `par`/`x0` the starting point, `fn`/`fun` the function to minimise and
`gr`/`jac` its gradient — the same ingredients our `grad_descent()` takes (this is
why we wrote everything functional-style: the interfaces match). Time to fit the
logistic regression from @sec-week5-losses, which has been waiting patiently for a
competent optimiser:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
fit_bfgs <- optim(par = c(0, 0, 0), fn = nll, gr = nll_grad, method = "BFGS")
fit_bfgs$par          # estimated coefficients
fit_bfgs$value        # NLL at the minimum
fit_bfgs$counts       # evaluations of fn and gr
fit_bfgs$convergence  # 0 means success
```

</div>

<div>

**Python**

```{python}
import scipy as sp

fit_bfgs = sp.optimize.minimize(nll, x0=np.zeros(3), jac=nll_grad, method="BFGS")
fit_bfgs.x            # estimated coefficients
fit_bfgs.fun          # NLL at the minimum
(fit_bfgs.nfev, fit_bfgs.njev)  # evaluations of fn and gr
fit_bfgs.success      # True means success
```

</div>
:::

**Note for Python users:** the report fields differ in name and convention: `$par`,
`$value`, `$counts` and `$convergence` become `.x`, `.fun`, `.nfev`/`.njev` and
`.success` — and R signals success with `convergence == 0`, scipy with
`success == True`. Always check it, whatever it is called.

About fifteen gradient evaluations in the R run (a dozen in the Python run, on its
own draw of the data). Let us line it up against our own algorithms — and against a
reference fit: `glm()`, R's official logistic regression fitter, on the left; on the
right, since Python has no built-in `glm()` (the usual tools live in the
`statsmodels` package), BFGS rerun with a very strict tolerance — measuring cost in
observation-gradients touched, as in @sec-week5-sgd:

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
# our GD, with a hand-tuned step size
path_nll <- grad_descent(nll, nll_grad, x0 = c(0, 0, 0), gamma = 2, n = 200)

# our SGD, with a hand-tuned decay schedule
nll_grad_i <- function(b, i) X_log[i, ] * (sigmoid(sum(X_log[i, ] * b)) - y_log[i])
set.seed(3)
path_sgd_nll <- sgd(nll_grad_i, c(0, 0, 0), gamma = \(t) 10 / (20 + t),
                    n_steps = 4000, n_data = n_log)

# the reference fit
fit_glm <- glm(y_log ~ X_log - 1, family = binomial)

results <- data.frame(
  method = c("glm()", "optim BFGS", "GD (200 steps)", "SGD (4000 steps)"),
  nll = c(nll(coef(fit_glm)), fit_bfgs$value,
          tail(path_nll$f, 1),
          nll(unlist(tail(path_sgd_nll, 1)[1:3]))),
  obs_gradients = c(NA, fit_bfgs$counts["gradient"] * n_log, 200 * n_log, 4000)
)
results$nll <- round(results$nll, 6)
results$obs_gradients <- format(results$obs_gradients, big.mark = ",", scientific = FALSE)
results
```

</div>

<div>

**Python**

```{python}
# our GD, with a hand-tuned step size
path_nll = grad_descent(nll, nll_grad, x0=np.zeros(3), gamma=2, n=200)

# our SGD, with a hand-tuned decay schedule
nll_grad_i = lambda b, i: X_log[i] * (sigmoid(X_log[i] @ b) - y_log[i])
rng = np.random.default_rng(3)
path_sgd_nll = sgd(nll_grad_i, np.zeros(3), gamma=lambda t: 10 / (20 + t),
                   n_steps=4000, n_data=n_log)

# the reference fit (no glm() here: BFGS again, with a very strict tolerance)
fit_ref = sp.optimize.minimize(nll, np.zeros(3), jac=nll_grad, method="BFGS",
                               options={"gtol": 1e-10})

results = pd.DataFrame({
    "method": ["reference (tight BFGS)", "minimize BFGS", "GD (200 steps)", "SGD (4000 steps)"],
    "nll": [fit_ref.fun, fit_bfgs.fun,
            path_nll["f"].iloc[-1],
            nll(path_sgd_nll.iloc[-1, :3].to_numpy())],
    "obs_gradients": [np.nan, fit_bfgs.njev * n_log, 200 * n_log, 4000]
})
results["nll"] = results["nll"].round(6)
results["obs_gradients"] = ["NA" if np.isnan(v) else f"{v:,.0f}"
                            for v in results["obs_gradients"]]
results
```

</div>
:::

Read the table slowly: it summarises the whole chapter. (The two columns fit two
different draws of the data, so the loss values differ between languages — the
*pattern* is the same.)

- **BFGS matches the reference fit — `glm()` in R — to six decimals** at the cost of
  a few thousand observation-gradients, with *no tuning at all* — no $\gamma$, no
  schedule, no $\beta$. The curvature information it accumulates replaces all the
  knobs we spent this chapter turning by hand.
- **GD gets there too**, but only after we hand-tuned $\gamma$, and at over ten
  times the cost. On an ill-conditioned problem (Exercise 5.4) the gap becomes a
  chasm.
- **SGD is nearly as accurate here at a fraction of the cost** — and its cost would
  be *unchanged* if $n$ were $10^8$, which is the regime it is built for.

Two closing remarks on `optim()` and `minimize()`. For problems with many parameters
there is `method = "L-BFGS-B"` (same name in both), a low-memory variant that stores
only the last few $(\mathbf{s}_k, \mathbf{y}_k)$ pairs instead of the full
$d \times d$ matrix (and accepts box constraints on the parameters). And practically:
**`optim()` with BFGS is what you reach for** when fitting a model by maximum
likelihood in R (`minimize()` in Python) — countless packages are, under the hood,
exactly `optim(par, nll, gr, method = "BFGS")` wearing a nice interface.

:::: {.callout-tip title="Exercise 5.4: optim in the wild"}
Fit a logistic regression predicting a car's transmission type from its horsepower
and weight, using the built-in `mtcars` data (`am` is 1 for manual; in Python the
same data frame ships with plotnine):

::: {style="display: grid; grid-template-columns: repeat(auto-fit, minmax(300px, 1fr)); grid-column-gap: 10px;"}
<div>

**R**

```{r}
#| eval: false
X_cars <- cbind(1, mtcars$hp, mtcars$wt)
y_cars <- mtcars$am
n_cars <- nrow(mtcars)

nll_cars <- \(b) {
  eta <- as.vector(X_cars %*% b)
  mean(log(1 + exp(eta)) - y_cars * eta)
}
nll_cars_grad <- \(b) {
  p <- 1 / (1 + exp(-as.vector(X_cars %*% b)))
  as.vector(t(X_cars) %*% (p - y_cars)) / n_cars
}
```

</div>

<div>

**Python**

```{python}
#| eval: false
from plotnine.data import mtcars

X_cars = np.column_stack([np.ones(len(mtcars)), mtcars["hp"], mtcars["wt"]])
y_cars = mtcars["am"].to_numpy()
n_cars = len(mtcars)

def nll_cars(b):
    eta = X_cars @ b
    return np.mean(np.log(1 + np.exp(eta)) - y_cars * eta)

def nll_cars_grad(b):
    p = 1 / (1 + np.exp(-(X_cars @ b)))
    return X_cars.T @ (p - y_cars) / n_cars
```

</div>
:::

a. Fit with `optim(c(0, 0, 0), nll_cars, nll_cars_grad, method = "BFGS")` and compare
   the NLL at the optimum with `glm(am ~ hp + wt, data = mtcars, family = binomial)`
   (compute `nll_cars(coef(...))`). Check `$convergence` — if it is not `0`, look up
   what that means in `?optim` and fix it via the `control` argument.
b. Now try your own `grad_descent()` with `gamma = 1e-3`, then with `gamma = 1e-4`
   and `n = 20000`. Describe what happens in each case.
c. Explain (b) in one or two sentences using @sec-week5-conditioning. Hint: look at
   the scales of the columns of `X_cars`. Which section of this chapter offers a fix
   that would let gradient descent compete?

**Note for Python users:** the fit in (a) is
`sp.optimize.minimize(nll_cars, np.zeros(3), jac=nll_cars_grad, method="BFGS")`; the
fields to check are `.success` and `.message`, and options such as iteration caps go
in `options = {"maxiter": 500}`. With no `glm()` in Python, compare against a
strict-tolerance refit (`options = {"gtol": 1e-10}`), as the reference row of the
table above did.
::::

## Where you will see all this again {#sec-week5-outro}

This chapter is a gateway. The same handful of ideas reappears, barely disguised,
all over statistics and machine learning:

- **Maximum likelihood, everywhere.** Almost every model you will fit from now on is
  fitted by numerically minimising a negative log-likelihood. Sometimes the optimiser
  is visible (`optim()` or `minimize()` in a package's source code), sometimes hidden
  (R's `glm()` uses iteratively reweighted least squares — a Newton variant), but it
  is always there, and now you know what its knobs mean and what its failure modes
  look like. When a model "fails to converge", you have seen the picture: a too-large
  step diverging, or an ill-conditioned likelihood crawling.
- **Neural networks.** Training a network is minimising an average loss over millions
  of observations and millions of parameters: mini-batch SGD with momentum is the
  core engine, and the popular Adam optimiser is momentum plus an adaptive *diagonal
  preconditioner* estimated on the fly. You have now implemented the essential parts
  of it in a few dozen lines of code.
- **Sampling algorithms.** When you meet Markov chain Monte Carlo in weeks 8–10 of
  this module, you will find the same geometry problem in a new costume: a sampler
  exploring an elongated, ill-conditioned posterior zig-zags just like gradient
  descent did, and the fix is again preconditioning (there it goes by names like
  "mass matrix" or "proposal covariance"). The valley picture from this chapter is
  worth keeping.

The tools were weeks 1–4: vectors and matrices to hold the data, loops to iterate,
functions to make the algorithms reusable, and plots to see what is really going on.
The ideas were this week's. Together they are most of computational statistics.
