# Bonus Week: Classes and Multiple Dispatch {#sec-bonus}

```{r}
#| include: false
set.seed(7432)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, out.width = "70%", fig.align = "center")
```

```{python}
#| include: false
import plotnine
plotnine.options.figure_size = (6, 4)
```

This chapter is **optional self-study material**: there is no timetabled slot for it, no lab,
and nothing here is examinable. It exists because at some point (probably while reading
someone else's package code) you will run into `setClass`, `@`, and `setMethod` in R, or
`class`, `self` and `__init__` in Python, and you deserve to know what is going on. It is
also, quietly, one of the most elegant corners of R.

Throughout the module we treated *functions* as the star of the show (@sec-week3). Here we
flip the perspective: instead of making functions smarter, we make the **data** smarter, by
attaching a *class* to it and letting the language pick the right behaviour automatically.

One warning about the shape of this chapter. Everywhere else in these notes the two
languages run side by side, line for line. Object-oriented programming is where they
genuinely part ways: R attaches behaviour to *generic functions* outside the object,
Python attaches it to *methods* inside the class, so a two-column view would be a lie.
Instead the chapter comes in two parts: @sec-bonus-r builds everything in R (S3, S4 and
multiple dispatch), and @sec-bonus-python rebuilds the same machinery in Python. Both
parts are for everyone: read the R part first, then the Python part; the contrast is the
real lesson.

## One name, many behaviours {#sec-bonus-motivation}

You have been using this machinery all term without noticing. Consider `summary()`:

```{r}
x <- rnorm(50)
summary(x)   # a five-number summary (plus the mean)
```

```{r}
fit <- lm(mpg ~ wt, data = mtcars)   # the mpg-vs-weight line from week 4
summary(fit)  # a full regression table!
```

Same function name, wildly different output. How does `summary()` know what to do? It asks
the object what it *is*:

```{r}
class(x)
class(fit)
```

`summary()` is a **generic function**: it does almost no work itself. It inspects the class
of its argument and *dispatches* to a specialised **method**: `summary.default()` for the
numeric vector, `summary.lm()` for the regression. There are many such methods lying around:

```{r}
head(methods("summary"), 8)
length(methods("summary"))
```

`print()` and `plot()` work exactly the same way: that is why typing `fit` at the console
shows a tidy coefficient display rather than the giant list that `fit` actually is. Contrast
this with `optim()` from @sec-week5, which returns a plain unclassed list and wishes you
luck.

**Note for Python:** the same phenomenon exists there: `len()` happily measures strings,
lists and arrays, and `+` adds numbers but concatenates strings. The mechanism, though, is
different: each type carries its own `__len__` or `__add__` method *inside* it.
@sec-bonus-python makes this precise.

This design is called **object-oriented programming** (OOP). R has several OOP systems; the
R part of this chapter meets the two that matter (S3 and S4) and finishes with S4's party
trick: dispatching on *two* arguments at once. Python has exactly one system, and we will
use it to rebuild the very same example.

## Classes in R {#sec-bonus-r}

Everything in this part is base R. The one sentence to hold on to, especially with the
Python part ahead: in R, methods belong to **generic functions**, not to objects. You will
write `summary(flips)`, never `flips.summary()`.

### S3: classes on a shoestring {#sec-bonus-s3}

S3 is R's oldest and simplest object system, and it powers most of what you have used so far
(`lm`, `data.frame`, `factor`...). The entire mechanism is two ingredients: a class
*attribute* on the object, and a naming convention (`generic.class`) for the methods.

#### The class attribute

Any object can be given a class (it is just a label):

```{r}
set.seed(7432)   # so we all flip the same coins

flips <- list(
  coin = "fair",
  prob = 0.5,
  outcomes = sample(c("H", "T"), 20, replace = TRUE)
)
class(flips) <- "coin_flips"
class(flips)
```

In practice we wrap this in a **constructor function** (just an ordinary function in the
sense of @sec-week3) so that all `coin_flips` objects are built the same way:

```{r}
coin_flips <- function(n, prob = 0.5, coin = "fair") {
  outcomes <- sample(c("H", "T"), n, replace = TRUE, prob = c(prob, 1 - prob))
  # structure() builds the object and attaches the class in one go
  structure(list(coin = coin, prob = prob, outcomes = outcomes),
            class = "coin_flips")
}

flips <- coin_flips(20)
```

If we print `flips` right now we get the raw list: R has no better idea. Let's fix that.

#### Writing methods for existing generics

To give our class its own printing behaviour, we simply define a function with the magic
name `print.coin_flips`. That naming convention *is* the registration:

```{r}
print.coin_flips <- function(x, ...) {
  cat("A", x$coin, "coin, flipped", length(x$outcomes), "times:\n")
  cat(paste(x$outcomes, collapse = ""), "\n")
  invisible(x)   # print methods return their input invisibly, by convention
}

flips
```

The `invisible(x)` follows the convention from @sec-week3: the method's real job is the
side effect (the `cat()`), so the return value should not print a second time.

A `summary` method works the same way (the first argument of `summary` is called `object`,
so we match that):

```{r}
summary.coin_flips <- function(object, ...) {
  n <- length(object$outcomes)
  heads <- sum(object$outcomes == "H")
  phat <- heads / n
  se <- sqrt(phat * (1 - phat) / n)
  cat("Flips:", n, "  Heads:", heads, sprintf("(%.0f%%)\n", 100 * phat))
  cat(sprintf("Approx. 95%% CI for P(H): [%.3f, %.3f]\n",
              max(0, phat - 1.96 * se), min(1, phat + 1.96 * se)))
  invisible(object)
}

summary(flips)
```

#### Writing your own generic

The generics themselves are one-liners: they call `UseMethod()`, which looks at the class of
the first argument and forwards the call to `generic.thatclass`, falling back to
`generic.default` if nothing more specific exists. Let's invent a generic `se()` for
standard errors:

```{r}
se <- function(x, ...) UseMethod("se")

# fallback: standard error of the mean of a plain numeric vector
se.default <- function(x, ...) sd(x) / sqrt(length(x))

# standard errors of regression coefficients
se.lm <- function(x, ...) sqrt(diag(vcov(x)))

# standard error of the estimated P(H)
se.coin_flips <- function(x, ...) {
  phat <- mean(x$outcomes == "H")
  sqrt(phat * (1 - phat) / length(x$outcomes))
}
```

One name, three behaviours:

```{r}
se(rnorm(100))
se(fit)
se(flips)
```

#### Inheritance: the class vector and NextMethod

The class attribute can be a *vector*, ordered from most to least specific. R tries
methods in that order. You have already met this in the wild:

```{r}
logit_fit <- glm(am ~ mpg, data = mtcars, family = binomial)
class(logit_fit)
```

A `glm` *is a* `lm` with extras: any generic without a `glm` method falls through to the
`lm` one. We can do the same, because a biased coin is still a coin:

```{r}
biased_flips <- function(n, prob) {
  out <- coin_flips(n, prob = prob, coin = "dodgy")
  class(out) <- c("biased_flips", class(out))   # new class FIRST, old one kept
  out
}

print.biased_flips <- function(x, ...) {
  cat("(!) This coin is loaded: P(H) =", x$prob, "\n")
  NextMethod()   # now run the next method in line: print.coin_flips
}

b <- biased_flips(20, prob = 0.9)
b
```

`NextMethod()` hands the call down the class vector, so the specialised method only adds
its extra bit and delegates the rest. And since we never wrote `summary.biased_flips`,
`summary(b)` falls through to `summary.coin_flips` automatically:

```{r}
inherits(b, "coin_flips")   # is-a check
summary(b)
```

That is S3 in full: a class attribute, `UseMethod`, a naming convention, and `NextMethod`.
No wonder most of CRAN runs on it.

::: {.callout-tip title="Exercise B.1: Print like a statistician"}
You are given this object:

```{r}
ci <- structure(
  list(estimate = 5.3, lower = 4.1, upper = 6.5, level = 0.95),
  class = "conf_int"
)
```

a. Write a `print.conf_int` method so that typing `ci` displays something like
   `95% CI: 5.30 [4.10, 6.50]` (use `sprintf()` for the formatting, and follow the
   invisible-return convention).
b. Check that `print(unclass(ci))` still shows the raw list: the data never changed,
   only the label R uses to choose a method.
c. Add a `se.conf_int` method for the `se()` generic above: assuming the interval is
   $\text{estimate} \pm 1.96 \times \text{se}$, recover the standard error as
   (upper − lower) / (2 × 1.96).
:::

### S4: classes with a seatbelt {#sec-bonus-s4}

S3 is charmingly informal. Nothing guarantees a `coin_flips` object actually contains
`outcomes`; worse, nothing stops us from simply lying:

```{r}
#| error: true
y <- 1:10
class(y) <- "lm"   # "trust me, I'm a regression"
summary(y)
```

R believed us, dispatched to `summary.lm`, and chaos ensued. **S4** (from the `methods`
package) fixes this with formal class definitions: declared slots with declared types,
optional validity checks, and formally registered generics and methods.

#### setClass, new and @

```{r}
library(methods)   # attached by default in normal sessions, but let's be explicit

setClass("Normal",
  slots = c(mean = "numeric", sd = "numeric"),
  prototype = prototype(mean = 0, sd = 1)   # default slot values
)
```

Objects are built with `new()`, and slots accessed with `@` (the S4 cousin of `$`):

```{r}
z <- new("Normal", mean = 1, sd = 2)
z@mean
new("Normal")    # prototype fills in the defaults: a standard normal
```

Unlike S3, the contract is enforced. A slot must exist and have the declared type:

```{r}
#| error: true
new("Normal", mean = "large")
```

#### Validity

Types are not enough: `sd = -3` is a perfectly good `numeric` and a terrible standard
deviation. A **validity function** encodes the statistical constraints; it returns `TRUE`
if all is well, or a string describing the problem:

```{r}
setClass("Normal",
  slots = c(mean = "numeric", sd = "numeric"),
  prototype = prototype(mean = 0, sd = 1),
  validity = function(object) {
    if (length(object@mean) != 1 || length(object@sd) != 1)
      return("mean and sd must be single numbers")
    if (object@sd <= 0)
      return("sd must be strictly positive")
    TRUE
  }
)
```

(You can also bolt a check onto an existing class later with `setValidity()`.) Now `new()`
polices the door:

```{r}
#| error: true
new("Normal", sd = -3)
```

One caveat: direct slot *assignment* does not re-run the validity function, so a valid
object can be corrupted after construction. `validObject()` performs the inspection on
demand:

```{r}
z <- new("Normal", mean = 1, sd = 2)
z@sd <- -3        # sneaks through...
```

```{r}
#| error: true
validObject(z)    # ...but does not survive an audit
```

#### Generics and methods, take two

In S4, generics are declared with `setGeneric()` and methods registered with `setMethod()`:
no naming conventions, everything explicit. Let's make a generic that evaluates a
distribution's density at some points:

```{r}
setGeneric("dens", function(d, x) standardGeneric("dens"))

setMethod("dens", "Normal", function(d, x) dnorm(x, mean = d@mean, sd = d@sd))

z <- new("Normal", mean = 1, sd = 2)
dens(z, 0:3)
```

#### show

The S4 analogue of the `print` method is `show()`, which controls what typing the object's
name displays:

```{r}
setMethod("show", "Normal", function(object) {
  cat("Normal(mean = ", object@mean, ", sd = ", object@sd, ")\n", sep = "")
})

z
```

#### Inheritance with contains

S4 inheritance is declared with `contains`: the subclass gets all the slots, the validity
check, and every method of its parent, plus whatever it adds:

```{r}
setClass("NamedNormal",
  contains = "Normal",
  slots = c(name = "character")
)

mu_prior <- new("NamedNormal", name = "prior for mu", mean = 0, sd = 10)
mu_prior@name
dens(mu_prior, 0)        # inherited method: a NamedNormal is a Normal
is(mu_prior, "Normal")
```

Try `new("NamedNormal", name = "oops", sd = -1)`: the parent's validity function fires
too. This is exactly how large systems like the `Matrix` package build families of matrix
classes (sparse, triangular, symmetric...) that all share the linear algebra you met in
@sec-week1.

::: {.callout-tip title="Exercise B.2: A Uniform with standards"}
Define an S4 class `Uniform` with numeric slots `min` and `max`, a prototype giving the
standard uniform on $[0, 1]$, and a validity function requiring both slots to be single
numbers with `min < max`. Then:

a. register a `show` method displaying something like `Uniform[0, 1]`;
b. register a `dens` method for the generic above (use `dunif()`);
c. confirm that `new("Uniform", min = 2, max = 1)` is rejected, and that
   `dens(new("Uniform"), c(0.5, 2))` returns `1 0`.
:::

### The highlight: multiple dispatch {#sec-bonus-multiple}

Here is where S4 earns its keep. `UseMethod` in S3 looks at the class of the **first
argument only**. That is fine for `summary(fit)`, but plenty of statistical operations
depend on the classes of *two* things at once.

The cleanest example is conjugate Bayesian updating. The posterior
$p(\theta \mid \mathbf{x}) \propto p(\mathbf{x} \mid \theta)\, p(\theta)$ has a closed form
only for matching *pairs*: a Beta prior with binomial data gives a Beta posterior, a Gamma
prior with Poisson data gives a Gamma posterior, and so on. The right computation depends on
**both** the prior's class **and** the data's class: a job for S4, which dispatches on the
full signature.

#### A tiny conjugate engine

First, the cast. We already have `Normal`; let's add two more priors and some data classes,
each data class carrying its model assumption in its name:

```{r}
# priors
setClass("Beta",  slots = c(a = "numeric", b = "numeric"))
setClass("Gamma", slots = c(shape = "numeric", rate = "numeric"))

# data
setClass("BinomData", slots = c(successes = "numeric", trials = "numeric"))
setClass("PoisData",  slots = c(x = "numeric"))
setClass("NormData",  slots = c(x = "numeric", sd = "numeric"))  # known noise sd
```

A quick cosmetic `show` for Beta (we leave the others to S4's serviceable default):

```{r}
setMethod("show", "Beta", function(object) {
  cat("Beta(", object@a, ", ", object@b, ")\n", sep = "")
})
```

Now the generic, with **two** dispatch arguments, and one method per conjugate pair. The
`signature()` names the class of each argument in turn:

```{r}
setGeneric("posterior", function(prior, data) standardGeneric("posterior"))

# Beta(a, b) + k successes in n trials  ->  Beta(a + k, b + n - k)
setMethod("posterior", signature("Beta", "BinomData"), function(prior, data) {
  new("Beta",
      a = prior@a + data@successes,
      b = prior@b + data@trials - data@successes)
})

# Gamma(shape, rate) + Poisson counts  ->  Gamma(shape + sum(x), rate + n)
setMethod("posterior", signature("Gamma", "PoisData"), function(prior, data) {
  new("Gamma",
      shape = prior@shape + sum(data@x),
      rate  = prior@rate + length(data@x))
})

# Normal prior + normal data with known sd  ->  Normal (precision-weighted average)
setMethod("posterior", signature("Normal", "NormData"), function(prior, data) {
  prec_prior <- 1 / prior@sd^2
  prec_data  <- length(data@x) / data@sd^2
  post_var   <- 1 / (prec_prior + prec_data)
  post_mean  <- post_var * (prec_prior * prior@mean + prec_data * mean(data@x))
  new("Normal", mean = post_mean, sd = sqrt(post_var))
})
```

Let's take it for a spin. A flat Beta prior, 7 successes in 20 trials:

```{r}
prior <- new("Beta", a = 1, b = 1)
d1 <- new("BinomData", successes = 7, trials = 20)
posterior(prior, d1)
```

The posterior is itself a `Beta`, so it can serve as the prior for the next batch of data.
Sequential updating comes for free, and matches pooling everything at once:

```{r}
d2 <- new("BinomData", successes = 5, trials = 10)
posterior(posterior(prior, d1), d2)
posterior(prior, new("BinomData", successes = 12, trials = 30))
```

The Normal-Normal pair reuses the `Normal` class from earlier (pretty `show`, validity
check and all):

```{r}
set.seed(7432)
obs <- rnorm(25, mean = 3, sd = 2)

posterior(new("Normal", mean = 0, sd = 10),
          new("NormData", x = obs, sd = 2))
```

#### The dispatch table

Conceptually, S4 keeps a **table** of methods indexed by the signature, here (prior
class, data class), and each call looks up the most specific matching entry:

|            | `BinomData`   | `PoisData`     | `NormData`      |
|------------|---------------|----------------|-----------------|
| `Beta`     | Beta update   |                |                 |
| `Gamma`    |               | Gamma update   |                 |
| `Normal`   |               |                | Normal update   |

Notice this is precisely the conjugacy table from your Bayesian statistics course, turned
into executable code. You can ask R to show its copy:

```{r}
showMethods("posterior")
```

#### An ANY fallback

What about the empty cells? `ANY` is the wildcard class, matched only when nothing more
specific fits: the S4 counterpart of `se.default`. We use it to fail *informatively*:

```{r}
setMethod("posterior", signature("ANY", "ANY"), function(prior, data) {
  stop("no conjugate update for a ", class(prior), " prior with ",
       class(data), " data -- time to write an MCMC sampler", call. = FALSE)
})
```

```{r}
#| error: true
posterior(new("Beta", a = 1, b = 1), new("PoisData", x = c(2, 0, 3)))
```

#### Prior to posterior, in pictures

A closing visual, with ggplot2 as in @sec-week4. We give `Beta` a `dens` method and plot
the flat prior against the posterior after 7 successes in 20 trials:

```{r}
library(ggplot2)

setMethod("dens", "Beta", function(d, x) dbeta(x, d@a, d@b))

post <- posterior(prior, d1)
grid <- seq(0.001, 0.999, length.out = 400)

df <- data.frame(
  theta   = rep(grid, 2),
  density = c(dens(prior, grid), dens(post, grid)),
  curve   = rep(c("prior", "posterior"), each = length(grid))
)

ggplot(df, aes(x = theta, y = density, colour = curve)) +
  geom_line(linewidth = 1) +
  labs(x = expression(theta), y = "density", colour = NULL) +
  theme_minimal()
```

Twenty coin flips took us from "no idea" to a tidy bump around $0.35$, and the machinery
that computed it never once asked us *which* update formula to use.

::: {.callout-tip title="Exercise B.3: One more row in the table"}
Exponential data with rate $\lambda$ is also conjugate to a Gamma prior:
$\text{Gamma}(\alpha, \beta)$ updated with observations $x_1, \dots, x_n$ gives
$\text{Gamma}(\alpha + n,\ \beta + \sum_i x_i)$.

a. Define an S4 class `ExpData` with a single numeric slot `x`.
b. Register a `posterior` method for the signature `("Gamma", "ExpData")`.
c. Simulate `set.seed(1); x <- rexp(50, rate = 2)` (Python:
   `rng = np.random.default_rng(1); x = rng.exponential(scale=1/2, size=50)`, because
   numpy parameterises the exponential by the *scale* $1/\lambda$, not the rate), update a
   `Gamma(1, 1)` prior with it, and check the posterior mean $\alpha/\beta$ lands within a
   couple of posterior standard deviations ($\sqrt{\alpha}/\beta$) of the true rate 2. The
   R and numpy streams share only the seed number, so the two languages will land on
   different values; both should pass the check.
d. Verify that updating with the first 25 observations and then the remaining 25 gives the
   same posterior as using all 50 at once (compare the slots with `all.equal()`; the
   floating-point lesson from @sec-week2 applies to Bayesians too).

(In Python, the same row fits the dict registry of @sec-bonus-python: add it there and
re-check c. and d.; the solutions notebook works it in both languages.)
:::

### The landscape, honestly {#sec-bonus-landscape}

R famously has *several* OOP systems. A field guide:

- **S3**: informal, everywhere. All of base R's modelling functions, and most of CRAN.
  If you write a package, you will almost certainly write S3 methods. Learn it properly.
- **S4**: formal, with declared slots, validity and multiple dispatch. Heavier ceremony,
  used where correctness matters at scale. As a statistician you will meet it mainly
  through the **Matrix** package (the sparse-matrix workhorse underneath `lme4`), the
  **Bioconductor** ecosystem (genomics: thousands of S4 classes), and the **distr**
  family of distribution packages, which does at production quality what our toy did here.
- **R5 / Reference Classes**: mutable objects whose methods live *inside* the object,
  Python-style; in practice superseded by the CRAN package **R6**, which does the same
  job leaner.
- **S7**: a successor to S3 and S4 being developed by the R Consortium, designed to
  unify the two; worth knowing the name, not yet worth reorganising your life around.

When in doubt: use S3 for everyday work, reach for S4 when you need validity guarantees or
genuinely need dispatch on more than one argument, and recognise the other two on sight.

For the definitive treatment, read the object-oriented programming chapters of Hadley
Wickham's *Advanced R* (2nd edition, free online at
[adv-r.hadley.nz](https://adv-r.hadley.nz)): chapter 13 for S3, chapter 15 for S4, and
chapter 16 for an honest comparison of the systems (a longer version of the advice above).

## Classes in Python {#sec-bonus-python}

If R's story is "several object systems, choose wisely", Python's is the opposite: there is
exactly **one**, it has been there from the start, and everything uses it. Every number,
string, array and function in Python is an object of some class. The deep difference from R
is *where behaviour lives*. In R it lives in generic functions outside the object
(`summary(flips)`); in Python it lives in methods inside the class, and you call them
*through* the object (`flips.summary()`). The rest of this part unpacks that one sentence,
and then rebuilds the conjugate-Bayes engine with it.

### `class`, `__init__` and `self`

A class is declared with `class`, and its constructor is a special method called
`__init__` (two underscores each side; Python calls these *dunder* methods). Its first
argument `self` is the object under construction; attaching data to `self` does in one
step what the R part did with `list(...)` plus a class attribute:

```{python}
import numpy as np
rng = np.random.default_rng(7432)   # the Python companion of set.seed(7432)

class CoinFlips:
    def __init__(self, n, prob=0.5, coin="fair"):
        self.coin = coin      # attributes are attached to self...
        self.prob = prob
        self.outcomes = rng.choice(["H", "T"], size=n, p=[prob, 1 - prob])

flips = CoinFlips(20)   # calling the class runs __init__
flips.coin              # ...and read back with a dot
type(flips)
```

**Note for R:** the class itself is the constructor: `CoinFlips(20)` does the job
of both the `coin_flips(20)` helper *and* `structure(..., class = ...)`. Attributes use a
dot where R's lists used `$` (and S4 used `@`), and there is no separate "attach the
label" step: an object's class is whichever class built it, forever queryable with
`type()`. One more thing you can see above: the flips will **not** match the R part's,
because `set.seed(7432)` and `default_rng(7432)` are different generators that merely
share a number. Expect a different H/T string; the statistics is the same.

### Methods and `__repr__`

Right now `flips` echoes as something like `<__main__.CoinFlips object at 0x...>`:
Python's version of R printing the raw list. Methods fix that. A method is just a function
defined inside the class body, taking `self` first; the special method `__repr__` is the
one the console uses to display an object, playing the role of `print.coin_flips` in S3
and of `show` in S4. Unlike R, where we bolted methods on one at a time from the outside,
a Python class is normally written in one block, so here is `CoinFlips` again, in full:

```{python}
class CoinFlips:
    def __init__(self, n, prob=0.5, coin="fair"):
        self.coin = coin
        self.prob = prob
        self.outcomes = rng.choice(["H", "T"], size=n, p=[prob, 1 - prob])

    def __repr__(self):
        return (f"A {self.coin} coin, flipped {len(self.outcomes)} times:\n"
                + "".join(self.outcomes))

    def summary(self):
        n = len(self.outcomes)
        heads = int(np.sum(self.outcomes == "H"))
        phat = heads / n
        se = np.sqrt(phat * (1 - phat) / n)
        print(f"Flips: {n}   Heads: {heads} ({100 * phat:.0f}%)")
        print(f"Approx. 95% CI for P(H): "
              f"[{max(0, phat - 1.96 * se):.3f}, {min(1, phat + 1.96 * se):.3f}]")

flips = CoinFlips(20)
flips                    # the console shows __repr__'s string
```

```{python}
flips.summary()          # methods are called through the object
```

Compare the shapes. R says `summary(flips)`: one generic, many methods, dispatch by class.
Python says `flips.summary()`: the object walks in already knowing its methods, and the
thing before the dot decides. It is the same dispatch decision, made at the opposite
address.

**Note for R:** no `invisible()` dance either: `summary` here has no `return`, so
it returns `None`, and `None` never echoes at the console. The printing *is* the job, and
that is the whole story.

### Inheritance and super()

Inheritance is declared in parentheses after the class name, and `super()` reaches the
parent class (R's `NextMethod`, but callable wherever you like):

```{python}
class BiasedFlips(CoinFlips):               # a BiasedFlips *is a* CoinFlips
    def __init__(self, n, prob):
        super().__init__(n, prob=prob, coin="dodgy")   # delegate construction to the parent

    def __repr__(self):
        return (f"(!) This coin is loaded: P(H) = {self.prob}\n"
                + super().__repr__())                  # our extra line, then the parent's display

b = BiasedFlips(20, prob=0.9)
b
```

```{python}
isinstance(b, CoinFlips)   # is-a check: the analogue of inherits()
b.summary()                # never defined on BiasedFlips -- inherited from CoinFlips
```

That is the whole inheritance story: `BiasedFlips` overrides what it must and inherits the
rest, exactly what the class vector plus `NextMethod` bought us in S3. Everyday Python,
mind you, often skips the formal `isinstance` question altogether: if an object has the
methods you need, use it; "if it quacks like a duck, it *is* a duck" (**duck typing**),
the informal spirit of S3 taken one cheerful step further.

### The same engine, one dict at a time {#sec-bonus-python-bayes}

Now the real test: can Python do the multiple-dispatch trick? Not natively: a method call
dispatches on the type of `self` and on nothing else. But the dispatch *table* at the heart
of the S4 machinery is just a lookup from a pair of classes to an update rule, and Python
has a perfectly good pair-to-anything lookup: a dictionary with tuple keys. We will build
the table ourselves.

The cast first: same priors, same data classes. `__repr__` does what our `show` methods
did, and `__init__` doubles as the validity check: there is no S4-style validator standing
at the door, so you write your own and `raise`:

```{python}
class Beta:
    def __init__(self, a, b):
        self.a = a
        self.b = b

    def __repr__(self):
        return f"Beta({self.a}, {self.b})"


class Gamma:
    def __init__(self, shape, rate):
        self.shape = shape
        self.rate = rate

    def __repr__(self):
        return f"Gamma(shape = {self.shape}, rate = {self.rate})"


class Normal:
    def __init__(self, mean=0, sd=1):   # defaults play the role of S4's prototype
        if sd <= 0:                     # __init__ doubles as the validity check
            raise ValueError("sd must be strictly positive")
        self.mean = mean
        self.sd = sd

    def __repr__(self):
        return f"Normal(mean = {self.mean}, sd = {self.sd})"
```

```{python}
class BinomData:
    def __init__(self, successes, trials):
        self.successes = successes
        self.trials = trials


class PoisData:
    def __init__(self, x):
        self.x = np.asarray(x)


class NormData:
    def __init__(self, x, sd):          # known noise sd
        self.x = np.asarray(x)
        self.sd = sd
```

Here is the multiple dispatch, spelled out by hand. The dictionary `conjugate` *is* the
dispatch table (its keys are `(prior class, data class)` pairs) and `posterior()` plays
the generic: it looks up the pair of types it was handed, with the `None` case as our
`ANY` fallback:

```{python}
conjugate = {}   # the dispatch table: (prior class, data class) -> update function

def posterior(prior, data):
    update = conjugate.get((type(prior), type(data)))
    if update is None:   # an empty cell of the table: fail informatively
        raise TypeError(f"no conjugate update for a {type(prior).__name__} prior with "
                        f"{type(data).__name__} data -- time to write an MCMC sampler")
    return update(prior, data)
```

Registering a method is now a dict assignment. Compare each entry with its `setMethod`
twin in @sec-bonus-multiple:

```{python}
# Beta(a, b) + k successes in n trials  ->  Beta(a + k, b + n - k)
def beta_binomial(prior, data):
    return Beta(prior.a + data.successes,
                prior.b + data.trials - data.successes)

conjugate[(Beta, BinomData)] = beta_binomial

# Gamma(shape, rate) + Poisson counts  ->  Gamma(shape + sum(x), rate + n)
def gamma_poisson(prior, data):
    return Gamma(prior.shape + np.sum(data.x),
                 prior.rate + len(data.x))

conjugate[(Gamma, PoisData)] = gamma_poisson

# Normal prior + normal data with known sd  ->  Normal (precision-weighted average)
def normal_normal(prior, data):
    prec_prior = 1 / prior.sd**2
    prec_data  = len(data.x) / data.sd**2
    post_var   = 1 / (prec_prior + prec_data)
    post_mean  = post_var * (prec_prior * prior.mean + prec_data * np.mean(data.x))
    return Normal(mean=post_mean, sd=np.sqrt(post_var))

conjugate[(Normal, NormData)] = normal_normal
```

The engine turns over exactly as before (same flat prior, same 7-in-20 data), and since
there is no randomness in the updates, the two languages agree to the digit:

```{python}
prior = Beta(1, 1)
d1 = BinomData(successes=7, trials=20)
posterior(prior, d1)
```

```{python}
d2 = BinomData(successes=5, trials=10)
posterior(posterior(prior, d1), d2)
posterior(prior, BinomData(successes=12, trials=30))
```

Sequential updating still comes for free and still matches the pooled batch. The
Normal-Normal pair, with a visible re-seed mirroring the R part:

```{python}
rng = np.random.default_rng(7432)   # fresh generator, as the R part re-called set.seed
obs = rng.normal(loc=3, scale=2, size=25)

posterior(Normal(mean=0, sd=10), NormData(x=obs, sd=2))
```

The draws differ from the R part's (again, shared seed number, different streams), so the
posterior mean will not match the R output; both runs, though, sit near the sample mean of
their own 25 draws, and the posterior sd collapses from the prior's 10 to just under 0.4.
(In this known-noise model the posterior sd does not depend on the draws at all, so *that*
number the two languages do agree on.)

Our fallback fails just as informatively as the `ANY` method did:

```{python}
#| error: true
posterior(Beta(1, 1), PoisData([2, 0, 3]))
```

And where R offered `showMethods("posterior")`, our table is an ordinary dict you can
simply look at:

```{python}
list(conjugate.keys())   # the filled cells of the conjugacy table
```

### Prior to posterior, in pictures (again)

The closing visual once more, now with plotnine, ggplot2's Python twin from the week 4
notes. Two Python-isms on the way: classes are themselves ordinary objects, so we can bolt
a density method onto `Beta` *after the fact* (a habit called *monkey-patching*, handy in
a notebook, frowned upon in production code), and it is `scipy.stats` playing the role of
R's `dbeta`:

```{python}
from scipy import stats
import pandas as pd
from plotnine import ggplot, aes, geom_line, labs, theme_minimal

# classes are objects too: bolt a density method onto Beta after the fact
Beta.dens = lambda self, x: stats.beta.pdf(x, self.a, self.b)

post = posterior(prior, d1)
grid = np.linspace(0.001, 0.999, 400)

df = pd.DataFrame({
    "theta":   np.concatenate([grid, grid]),
    "density": np.concatenate([prior.dens(grid), post.dens(grid)]),
    "curve":   np.repeat(["prior", "posterior"], len(grid)),
})

(ggplot(df, aes(x="theta", y="density", colour="curve"))
 + geom_line(size=1)                  # plotnine still calls linewidth "size"
 + labs(x="θ", y="density", colour="")   # no expression() needed: just type θ
 + theme_minimal()).show()
```

Same twenty flips, same tidy bump around $0.35$: the two languages disagree about where
behaviour should live, but not about the posterior.

### Two tables, one honest comparison {#sec-bonus-compare}

Step back and compare what you just did twice.

In R, S4 made the dispatch table *the language's problem*: `setMethod` filed each update
under its pair of classes, and lookup, inheritance-aware matching (a `NamedNormal` prior
finds the `Normal` method automatically) and the `ANY` fallback all came built in. In
Python, dispatch on more than one argument is *your problem*: the built-in mechanism
(choosing behaviour by the thing before the dot) is **single** dispatch on `self`'s type,
and the standard library's `functools.singledispatch` automates single dispatch for plain
functions but stops there too. So we built the pair table by hand: a dict, a two-line
lookup, a `raise`. That is precisely the machinery S4 automates, and our hand-rolled
version is honestly dumber, too: `type(prior)` matches exactly, so a *subclass* of `Beta`
would sail straight past the `(Beta, BinomData)` entry that S4's inheritance-aware
dispatch would have found. Making the dict version subclass-aware is a nice puzzle, not a
one-liner.

The trade cuts both ways. Python's single system is small, universal, and never leaves you
wondering which of four object frameworks a package is written in. R's S4 is heavier
ceremony, but it puts a genuinely statistical kind of dispatch (behaviour that depends on
a prior *and* a likelihood at once) directly into the language. Now you know what you are
looking at when you meet either.
